# Basics 

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter covers the very basics of Python and R.

## Math operations

Mathematical operators are the same except for exponents, integer division, and remainder division (modulo).

#### Python {-}

Python uses `**` for exponentiation, `//` for integer division, and `%` for remainder division.

```{python}
3**2
5 // 2
5 % 2
```

In Python, the `+` operator can also be used to combine strings. See this TBD section.

#### R {-}

Python uses `^` for exponentiation, `%/%` for integer division, and `%%` for remainder division.


```{r}
3^2
5 %/% 2
5 %% 2
```

## Assignment

Python uses `=` for assignment while R can use either `=` or `<-` for assignment. The latter "assignment arrow" is preferred in most R style guides to distinguish it between assignment and setting the value of a function argument. According to R's documentation, "The operator `<-` can be used anywhere, whereas the operator `=` is only allowed at the top level (e.g., in the complete expression typed at the command prompt) or as one of the subexpressions in a braced list of expressions." See `?assignOps`.

#### Python {-}

```{python}
x = 12
```

#### R {-}

```{r}
x <- 12
```

## Printing a value

To see the value of an object created via assignment, you can simply enter the object at the console and hit enter for both Python and R, though it is common in Python to explicitly use the `print()` function.

#### Python {-}

```{python}
x
```

#### R {-}

```{r}
x
```


## Packages and Libraries

User-created functions can be bundled and distributed as libraries (Python) and packages (R). Libraries and packages need to be installed only once. Thereafter, they're "imported" (Python) or "loaded" (R) in each new session when needed.

Libraries and packages with large user bases are often updated to add functionality and fix bugs. The updates are not automatically installed. Staying apprised of library/package updates can be challenging. Some suggestions for doing so are following developers on Twitter, signing up for newsletters, or periodically checking to see what updates are available.

Libraries and packages often depend on other libraries and packages. These are known as "dependencies." Sometimes, libraries/packages are updated to accommodate changes to libraries/packages they depend on.

#### Python {-}


#### R {-}

The main repository for R packages is the [Comprehensive R Archive Network](https://cran.r-project.org/) (CRAN). Another repository is [Bioconductor](https://www.bioconductor.org/), which provides tools for working with genomic data. Many packages are also distributed on [GitHub](https://github.com/).

To install packages from CRAN, use the `install.packages()` function. In RStudio, you can also go to Tools...Install Packages... for a dialog that will auto-complete package names as you type.

```{r eval=FALSE}
# install the vcd package, a package for Visualizing Categorical Data
install.packages("vcd")

# load the package
library(vcd)

# see which packages on your computer have updates available
old.packages()

# download and install available package updates;
# set ask = TRUE to verify installation of each package
update.packages(ask = FALSE)
```

To install R packages from GitHub use the `install_github()` function from the **devtools** package. You need to include the username of the repo owner followed by a forward slash and the name of the package. Typing two colons between a package and a function in the package allows you to use that function without loading the package.

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("username/packagename")
```

Occasionally, when installing or updating a package you will be asked "Do you want to install from sources the package which needs compilation?" R packages on CRAN are _compiled_ for Mac and Windows operating systems. That can take a day or two. If you try to install a package that has not been compiled then you'll get asked if you want to compile it yourself. You can usually just answer No. If you answer Yes, R will try to compile the package on your computer. This will only work if you have the required build tools on your computer. For Windows this means having [RTools](https://cran.r-project.org/bin/windows/Rtools/) installed. Mac users typically already have the necessary build tools.

## Generating a sequence of values

There are multiple ways in both Python and R to generate sequences of values. In Python, one option is `arange()` from **NumPy**. In R, a common approach is to use `seq()`. Sequences can be incremented by indicating a `step` argument in `arange()` or a `by` argument in `seq()`. Be aware that the start/stop interval in `arange()` is _open_, but the from/to interval in `seq()` is _closed_.

#### Python {-}

```{python}
import numpy as np
x = np.arange(start = 1, stop = 11, step = 2)
print(x)
```

#### R {-}

```{r}
x <- seq(from = 1, to = 11, by = 2)
x
```

## Calculating means and medians

The **NumPy** Python library has functions for calculating means and medians, and base R has functions for doing the same.

#### Python {-}

```{python}
# Mean, using function from NumPy library
import numpy as np
x = [90, 105, 110]
x_avg = np.mean(x)
print(x_avg)

# Median, using function from NumPy library
x = [98, 102, 20, 22, 304]
x_med = np.median(x)
print(x_med)
```

#### R {-}

```{r}
# Mean, using function from base R
x <- c(90, 105, 110)
x_avg <- mean(x)
x_avg

# Median, using function from base R
x <- c(98, 102, 20, 22, 304)
x_med <- median(x)
x_med
```