# Basics 

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter covers the very basics of Python and R.

## Math

Mathematical operators are the same except for exponents, integer division, and remainder division (modulo).

#### Python {-}

```{python}
3**2
5 // 2
5 % 2
```

#### R {-}

```{r}
3^2
5 %/% 2
5 %% 2
```

In Python, the `+` operator can also be used to combine strings. See this TBD section.

## Assignment

Python uses `=` for assignment while R can use either `=` or `<-` for assignment. The latter "assignment arrow" is preferred in most R style guides to distinguish it between assignment and setting the value of a function argument. According to R's documentation, "The operator `<-` can be used anywhere, whereas the operator `=` is only allowed at the top level (e.g., in the complete expression typed at the command prompt) or as one of the subexpressions in a braced list of expressions." See `?assignOps`.

#### Python {-}

```{python}
x = 12
```

#### R {-}

```{r}
x <- 12
```

## Printing a value

To see the value of an object created via assignment, you can simply enter the object at the console and hit enter for both Python and R, though it is common in Python to explicitly use the `print` function.

#### Python {-}

```{python}
x
```

#### R {-}

```{r}
x
```

## Generating a sequence of values

There are multiple ways in both Python and R to generate sequences of values. In Python, one option is `arange()` from `numpy`. In R, a common approach is to use `seq()`. The sequences can be incremented by indicating a `step` argument in `arange()` or a `by` argument in `seq()`. Be aware that the start/stop interval in `arange()` is _open_, but the from/to interval in `seq()` is _closed_.

#### Python {-}

```{python}
import numpy as np
x = np.arange(start = 1, stop = 11, step = 2)
x
```

#### R {-}

```{r}
x <- seq(from = 1, to = 11, by = 2)
x
```