# Data Structures 

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter compares and contrasts data structures in Python and R.

## One-dimensional data

A one-dimensional data structure can be visualized as a column in a spreadsheet or as a list of values. 

#### Python {-}


#### R {-}

In R a one-dimensional data structure is called a _vector_. We can create a vector using the `c()` function. A vector in R can only contain one type of data (all numbers, all strings, etc). The columns of data frames are vectors. If multiple types of data are put into a vector, the data will be coerced according to the hierarchy `logical` < `integer` < `double` < `complex` < `character`. This means if you mix, say integers and character data, all the data will be coerced to character. 

```{r}
x1 <- c(23, 43, 55)
x1

# all values coerced to character
x2 <- c(23, 43, 'hi')
x2
```

Values in a vector can be accessed by position using indexing brackets. 

```{r}
# extract the 2nd value
x1[2]

# extract the 2nd and 3rd value
x1[2:3]

```


## Two-dimensional

Two-dimensional data is rectangular in nature, consisting of rows and columns. These can be the type of data you might find in a spreadsheet with a mix of data types in columns, or matrices as you might encounter in matrix algebra.

#### Python {-}


#### R {-}

Two-dimensional data structures in R include the _matrix_ and _data frame_. A matrix can contain only one data type. A data frame can contain multiple vectors each of which can consist of different data types. 

Create a matrix with the `matrix()` function. Create a data frame with the `data.frame()` function. Most imported data comes into R as a data frame.

```{r}
# matrix
m <- matrix(data = c(1,3,5,7), nrow = 2, ncol = 2)
m

# data frame
d <- data.frame(name = c("Rob", "Cindy"),
                age = c(35, 37))
d
```

Values in a matrix and data frame can be accessed by position using indexing brackets. The first number(s) refers to rows, the second number(s) to columns. Leaving row or column numbers empty selects all rows or columns.

```{r}
# extract value in row 1, column 2
m[1,2]

# extract values in row 2
d[2,]
```

