# Importing Data 

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter reviews importing external data into Python and R, including CSV, txt, and other structured data files. 

## CSV

Comma separated value (CSV) files are text files with fields separated by commas. They are useful for "rectangular" data where rows represent observations and columns represent variables or features. 

#### Python {-}



```{python}
import pandas
d = pandas.read_csv('data/dat.csv')
```


#### R {-}

There are many ways to import a csv file. A common way is to use the base R function `read.csv()`.

```{r eval=FALSE}
d <- read.csv("data/dat.csv")

```

Two packages that provide alternatives to `read.csv()` are **readr** and **data.table**. The **readr** function `read_csv()` returns a [tibble](https://r4ds.had.co.nz/tibbles.html). The **data.table** function `fread()` returns a [data.table](https://rdatatable.gitlab.io/data.table/articles/datatable-intro.html). 

```{r eval=FALSE, message=FALSE}
library(readr)
d <- read_csv("data/dat.csv")

library(data.table)
d <- fread("data/dat.csv")
```


## XLS/XLSX (Excel)


#### Python {-}

#### R {-}


## Text 


#### Python {-}

#### R {-}


## JSON

#### Python {-}

#### R {-}


## XML

#### Python {-}

#### R {-}


