# Data Manipulation

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at various strategies for filtering, selecting, modifying and deriving variables in data. Unless otherwise stated, examples are for DataFrames (Python) and data frames (R) and use the mtcars data frame that is included with R.

```{python}
# Python
import pandas
mtcars = pandas.read_csv('data/mtcars.csv')
```


```{r}
# R
data(mtcars)
# drop row names to match Python version of data
rownames(mtcars) <- NULL
```


## Names of variables and their types

View and inspect the names of variables and their type (numeric, string, logical, etc.) This is useful to ensure that variables have the expected type. 

#### Python {-}

The `.info()` function in pandas lists information on the DataFrame.


Setting the argument `verbose` to `True` prints the name of the columns, their length excluding `NULL` values, and their data type (`dtype`) in a table. The function lists the unique data types in the DataFrame, and it prints how much memory the DataFrame takes up.

```{python}
mtcars.info(verbose=True)
```

Setting `verbose` to `False` excludes the table describing each column.

```{python}
mtcars.info(verbose=False)
```

If a DataFrame has 100 or fewer columns, the `verbose` argument defaults to `True`. 

#### R {-}

The `str()` function in R lists the names of the variables, their type, the first few values, and the dimensions of the data frame.  

```{r}
str(mtcars)
```

To see just the names of the data frame, use the `names()` function.

```{r}
names(mtcars)
```

To see just the dimensions of the data frame, use the `dim()` function. It returns the number of rows and columns, respectively.


```{r}
dim(mtcars)
```

## Select variables

How to select specific columns of data frames.

#### Python {-}
The period operator `.` provides access to a column in a DataFrame as a vector. This returns pandas Series. A pandas series can do everything a numpy array can do.

```{python}
mtcars.mpg
```

Indexing also provides access to columns as a pandas Series. Single and double quotations both work.

```{python}
mtcars['mpg']
```

Operations on numpy arrays are faster than operations on pandas series. But using pandas series should be fine, in terms of performance, in many cases. This is important for large data sets on which many operations are performed. The `.values` function returns a numpy array.

```{python}
mtcars['mpg'].values
```

Double indexing returns a pandas DataFrame, instead of a numpy array or pandas series.

```{python}
mtcars[['mpg']]
```

The `head()` and `tail()` functions return the first 5 or last 5 values. Use the `n` argument to change the number of values. This function works on numpy arrays, pandas series and pandas DataFrames.

```{python}
# first 6 values
mtcars.mpg.head()
```

```{python}
# last row of DataFrame
mtcars.tail(n=1)
```

#### R {-}

The dollar sign operator, `$`, provides access to a column in a data frame as a vector.

```{r}
mtcars$mpg
```

Double-indexing brackets also provide access to columns as a vector.

```{r}
mtcars[["mpg"]]
```

Single-indexing brackets work as well, but they return a data frame instead of a vector (if used with a data frame).

```{r}
mtcars["mpg"]
```


Single-indexing brackets also allow selection of rows when used with a comma. The syntax is `rows, columns`

```{r}
# first three rows
mtcars[1:3, "mpg"]
```

Finally single-indexing brackets allow us to select multiple columns. Request columns either by name or position using a vector.

```{r}
mtcars[c("mpg", "cyl")] 
# same as mtcars[1:2] 
```

The `head()` and `tail()` functions return the first 6 or last 6 values. Use the `n` argument to change the number of values. They work with vectors or data frames.

```{r}
# first 6 values
head(mtcars$mpg)
```

```{r}
# last row of data frame
tail(mtcars, n = 1)
```


## Filter/Subset variables

How to view rows of a data frame that meet certain conditions.

#### Python {-}

We can filter rows of a data frame based on a condition to subset on.

```{python}
mtcars[mtcars["mpg"] > 30]
```

To select particular columns with the condition, subset the data frame using a list of desired columns as indices.

```{python}
mtcars[mtcars["mpg"] > 30][["mpg", "cyl"]]
```

You can also filter with multiple row conditions.

```{python}
mtcars[mtcars["mpg"] > 30][mtcars["hp"] < 66][["mpg", "cyl"]]
```

#### R {-}

In base R, we can use subsetting brackets or the `subset()` function to select rows based on some condition. Below we demonstrate both approaches to view only those rows with "mpg" greater than 30. First we begin with subsetting brackets.

The subsetting brackets take three arguments: 

1. `i`: the condition to subset on.
2. `j`: the columns to show. If none specified, all columns are returned
3. `drop`: an optional logical argument (TRUE/FALSE) to determine whether or not to coerce the output to the lowest possible dimension. The default is TRUE.

We rarely type the first two argument names, `i` and `j`, when using subsetting brackets.

This example returns only the rows with mpg > 30 and all columns. Notice we need to preface mpg with `mtcars$` to tell R where to find the "mpg" column and that we need to provide a comma after the condition.

```{r}
mtcars[mtcars$mpg > 30, ]
```

We can select what columns to see in the second argument as a vector. Notice we only need to specify the column names as a character vector. We can also use numbers corresponding to the column number as well as conditional statements.

```{r}
mtcars[mtcars$mpg > 30, c("mpg", "wt", "gear")]
```

Show first three columns.

```{r}
mtcars[mtcars$mpg > 30, 1:3]
```

Show columns with names consisting of only two characters. The `nchar()` function counts the number of characters in a string. The expression `nchar(names(mtcars)) == 2` returns a vector of TRUE/FALSE values where TRUE indicates the column name is only two characters in length.

```{r}
mtcars[mtcars$mpg > 30, nchar(names(mtcars)) == 2]
```

Notice when we specify only one column, the brackets return a vector.

```{r}
mtcars[mtcars$mpg > 30, "mpg"]
```

To get a data frame, set the `drop` argument to FALSE.

```{r}
mtcars[mtcars$mpg > 30, "mpg", drop = FALSE]
```

The `subset()` function allows us to refer to column names without using the `$` extractor function or quoting column names. It also has a drop argument but its default is FALSE. It has four arguments:

1. `x`: the data frame to subset.
2. `subset`: the condition to subset on.
3. `select`: the columns to select.
4. `drop`: an optional logical argument (TRUE/FALSE) to determine whether or not to coerce the output to the lowest possible dimension. The default is FALSE.

We rarely type the first three argument names, `x`, `subset` and `select`, when using `subset()`.

Below we replicate the previous examples using `subset()`.

```{r}
# rows where mpg > 30 and all columns
subset(mtcars, mpg > 30)
```

```{r}
# rows where mpg > 30 and the mpg, wt, and gear columns
subset(mtcars, mpg > 30, c(mpg, wt, gear))
```

```{r}
# rows where mpg > 30 and the first three columns
subset(mtcars, mpg > 30, 1:3)
```

```{r}
# rows where mpg > 30 and columns consisting of two characters
subset(mtcars, mpg > 30, nchar(names(mtcars)) == 2)
```

```{r}
# rows where mpg > 30 and mpg column, as a vector
subset(mtcars, mpg > 30, mpg, drop = TRUE)
```

```{r}
# rows where mpg > 30 and mpg column, as a data frame
subset(mtcars, mpg > 30, mpg)
```

Another difference between subsetting brackets and the `subset()` function is how they handle missing values. Subsetting brackets return missing values while `subset()` does not. We demonstrate with a toy data frame. Notice the "x" column has a missing value.

```{r}
dframe <- data.frame(x = c(1, NA, 5), 
                     y = c(12, 21, 34))
dframe
```

When we condition on x < 3, the subsetting bracket approach returns a row with NA values.

```{r}
dframe[dframe$x < 3,]
```

The `subset()` approach ignores the missing value.

```{r}
subset(dframe, x < 3)
```

To replicate the `subset()` result with the subsetting brackets, we need to include an additional condition to only show rows where x is NOT missing. We can do that with the `is.na()` function. The `is.na()` function returns TRUE if a value is missing and FALSE otherwise. If we preface with `!`, we get TRUE if a value is NOT missing and FALSE otherwise. 

```{r}
dframe[dframe$x < 3 & !is.na(dframe$x),]
```

See also the `filter()` function in the [**dplyr**](https://dplyr.tidyverse.org/) package and the enhanced subsetting brackets in the [**data.table**](https://rdatatable.gitlab.io/data.table/) package.

## Rename variables

How to rename variables or "column headers".

#### Python {-}

Column names can be changed using the function `.rename()`. Below, we change the column names "cyl" and "wt" to "cylinder" and "WT", respectively.

```{python}
mtcars.rename(columns={"cyl":"cylinder", "wt":"WT"})
```

Alternatively, column names can be changed by replacing the vector of column names with a new vector. Below, we create a vector of columns that replaces "drat" with "axle_ratio" using conditional match and indexing and "disp" with "DISP" using indexing.

```{python}
column_names = mtcars.columns.values

# using conditional match
column_names[column_names == "drat"] = "axle_ratio"

# using indexing
column_names[2] = "DISP"

mtcars.columns = column_names
mtcars.columns
```



#### R {-}

Variable names can be changed by their index (ie, order of columns in the data frame). Below the second column is "cyl". We change the name to "cylinders".

```{r}
names(mtcars)[2]
names(mtcars)[2] <- "cylinders"
names(mtcars)
```

Variable names can also be changed by conditional match. Below we find the variable name that matches "drat" and change to "axle_ratio".

```{r}
names(mtcars)[names(mtcars) == "drat"]
names(mtcars)[names(mtcars) == "drat"] <- "axle_ratio"
names(mtcars)
```

More than one variable name can be changed using a vector of positions or matches.

```{r}
names(mtcars)[c(6,8)] <- c("weight", "engine")

# or
# names(mtcars)[names(mtcars) %in% c("wt", "vs")] <- c("weight", "engine")

names(mtcars)
```

See also the `rename()` function in the [**dplyr**](https://dplyr.tidyverse.org/). 

## Create, replace and remove variables

We often need to create variables that are functions of other variables, or replace existing variables with an updated version.

#### Python {-}
Adding a new variable using the indexing notation and assigning a result adds a new column.

```{python}
# add column for Kilometer per liter
mtcars['kpl'] = mtcars.mpg/2.352
```

Doing the same with an _existing_ column name updates the values in a column.

```{python}
# update to liters per 100 Kilometers
mtcars['kpl'] = 100/mtcars.kpl 
```

Alternatively, the `.` notation can be used to update the values in a column.
```{python}
# update to liters per 50 Kilometers
mtcars.kpl = 50/mtcars.kpl 
```

To remove a column, use the `.drop()` function.

```{python}
# drop the kpl variable
mtcars.drop(columns=['kpl'])
```


#### R {-}

Adding a new variable name after the dollar sign notation and assigning a result adds a new column.

```{r}
# add column for Kilometer per liter
mtcars$kpl <- mtcars$mpg/2.352
```

Doing the same with an _existing_ variable updates the values in a column. 

```{r}
# update to liters per 100 Kilometers
mtcars$kpl <- 100/mtcars$kpl 
```

To remove a variable, assign it `NULL`.

```{r}
# drop the kpl variable
mtcars$kpl <- NULL
```

See also the `mutate()` function in the [**dplyr**](https://dplyr.tidyverse.org/) package.

## Create strings from numbers

You may have data that is numeric but that needs to be treated as a string. 

#### Python {-}
You can change the data type of a column in a DataFrame using the `astype` function.

```{python}
mtcars['am'] = mtcars['am'].astype(str)
type(mtcars.am[0]) # check the type of the first item in 'am' column
```

A potential number-to-string conversion task in Python might be formatting 5-digit American zip codes. Some zip codes begin with 0, but if stored as a numeric value, the 0 is dropped. For example, consider the following pandas DataFrame. Notice the leading 0 is dropped from two of the zip codes.

```{python}
zc = pandas.read_csv('data/zc.csv')
print(zc)
```

One way to fix this is using the string `zfill()` method. First we convert the numeric column to string type using the method we just demonstrated. Then we access the "zip" column using `zc.zip` and the `zfill()` method using `str.zfill` with the width parameter set to 5. This pads the string with "0" on the left to make each value 5 characters wide. 

```{python}
zc['zip'] = zc['zip'].astype(str)
zc['zip'] = zc.zip.str.zfill(5)
print(zc)
```

If we knew we were importing zip codes using `read_csv`, we could also use the `dtype` argument to specify which storage type to use for the "zip" column. Below we pass a dictionary that maps the "str" type to the "zip" column. The result is a properly formatted zip code column.

```{python}
zc = pandas.read_csv('data/zc.csv', dtype = {'zip': 'str'})
print(zc)
```


#### R {-}

The `as.character()` function takes a vector and converts it to string format.

```{r}
head(mtcars$am)
head(as.character(mtcars$am))
```

Note we just demonstrated conversion. To save the conversion we need to _assign_ the result to the data frame.

```{r}
# add new string variable am_ch
mtcars$am_ch <- as.character(mtcars$am)
head(mtcars$am_ch)
```


The `factor()` function can also be used to convert a numeric vector into a  categorical variable. The result is not exactly a string, however. A factor is made of integers with character labels. Factors are useful for character data that have a fixed set of levels (eg, "grade 1", grade 2", etc)

```{r}
# convert to factor
head(mtcars$am)
head(factor(mtcars$am))

# convert to factor with labels
head(factor(mtcars$am, labels = c("automatic", "manual")))
```

Again we just demonstrated factor conversion. To save the conversion we need to assign to the data frame.

```{r}
# create factor variable am_fac
mtcars$am_fac <- factor(mtcars$am, labels = c("automatic", "manual"))
head(mtcars$am_fac)
```

A common number-to-string conversion task in R is formatting 5-digit American zip codes. Some zip codes begin with 0, but if stored as a numeric value, the 0 is dropped. 

```{r}
zip_codes <- c(03766, 03748, 22901, 03264)
zip_codes
```

We need to store the zip code as a character value so the 0 is preserved. One way to do this is via the `sprintf()` function in base R. The first argument is the _format string_ or _conversion specification_. A conversion specification begins with "%". The following "0" and "5" says to format the zip_codes vector as a 5-digit string padded by zeroes on the left. The final "i" says we're working with integer values. 

```{r}
sprintf("%05i", zip_codes)
```

See also the `str_pad()` function in the **stringr** package.


## Create numbers from strings

String variables that ought to be numbers usually have some character data in the values such as units (eg, "4 cm"). To create numbers from strings it's important to remove any character data that cannot be converted to a number.

#### Python {-}
The `astype(float)` or `astype(int)` function will coerce strings to numerical representation.

For demonstration, let's say we have the following numpy array.

```{python}
import numpy as np
weight = np.array(["125 lbs.", "132 lbs.", "156 lbs."])
```

The `astype(float)` function throws an error due to the presence of strings. The `astype()` function is for numpy arrays.

```{python}
try:
  weight.astype(float)
except ValueError:
  print("ValueError: could not convert string to float: '125 lbs.'")
```

One way to approach this is to first remove the strings from the objects and then use `astype(float)`. Below we use the `strip()` function to find " lbs." using a list comprehension.

```{python}
# [] indicates a list in python
# np.array() changes the list back into an array
weight = np.array([w.strip(" lbs.") for w in weight])
```

Now we can use the `astype()` function to change the elements in weight from `str` to `float`.

```{python}
weight.astype(float)
```

#### R {-}

The `as.numeric()` function will attempt to coerce strings to numeric type _if possible_. Any non-numeric values are coerced to NA.

For demonstration, let's say we have the following vector.

```{r}
weight <- c("125 lbs.", "132 lbs.", "156 lbs.")
```

The `as.numeric()` function returns all NA due to presence of character data.

```{r}
as.numeric(weight)
```

There are many ways to approach this. A common approach is to first remove the characters and then use `as.numeric()`. Below we use the `gsub()` function to find "lbs." and replace with nothing (find-and-replace procedures are discussed more below).

```{r}
weightN <- gsub("lbs.", "", weight)
as.numeric(weightN)
```

The `parse_number()` function in the **readr** package can often take care of these situations automatically.

```{r}
readr::parse_number(weight)
```

## Combine strings

String concatenation---turning 'Jane' and 'Smith' into 'Jane Smith'---is easily done in both languages.

#### Python {-}

The `+` operator can combine strings in Python.

```{python}
species = 'yellow-bellied sea snake'
tail_shape = 'paddle-shaped'

statement = 'The ' + species + ' has a ' + tail_shape + ' tail that helps it swim.'
print(statement)
```

#### R {-}

The `paste()` and `paste0()` functions combine strings in R. The former concatenates strings and places spaces between them; the latter concatenates sans spaces.

```{r}
species <- 'rainbow boa'
appearance <- 'iridescent'
location <- 'Central and South America'

statement1 <- paste('The', species, 'has an', appearance, 'sheen.')
statement1

# Note that spaces must be provided explicitly when using paste0()
statement2 <- paste0('The ', species, ' is found in ', location)
statement2
```

## Finding and replacing patterns within strings

This section reviews key functions in Python and R for finding and replacing character patterns. The functions we discuss can search for fixed character patterns (e.g., "Meredith Rollins" to case-sensitively match that name and that name alone) or regular expression (regex) patterns (e.g., `\w+` to capture all instances of >=1 word character). Note that in R, meta characters, like `w` (to match word characters) and `d` (to match digits), are escaped with _two_ backslashes (e.g., `\\w` and `\\d`). In Python, regex patterns are generally headed by `r`, which allows meta characters in the regex itself to be escaped with just one `\` (e.g., `r"\w+"`). Regex is an enormous topic, and we don't discuss it at any length here, but you can learn more about regular expressions---and how they're implemented in different programming languages---at these resources: https://www.regular-expressions.info/; https://regexone.com/

#### Python {-}

The **re** module provides a set of functions for searching and manipulating strings. The `search()` function does exactly as its name suggests: It identifies matches for a fixed or regex character pattern in a string. `sub()` searches for and replaces character patterns (fixed or regex). The `count` argument in `sub()` allows a user to specify how many instances of the matched pattern they want to to replace; e.g., use `count = 1` to replace just the first instance of a match.

```{python}
import re
statement = 'Pencils with an HB graphite grade are commonly used for writing. An HB pencil is approximately equal to a #2 pencil.'

# Search for "HB" using fixed and regex patterns
search_result1 = re.search(pattern = "HB", string = statement)
print(search_result1)
search_result2 = re.search(pattern = r"[H,B]{2}", string = statement)
print(search_result2)

# Replace all instances of "HB"
all_replaced = re.sub(pattern = 'HB', repl = 'HB (hard black)', string = statement)
print(all_replaced)

# Replace just the first instance of HB
one_replaced = re.sub(pattern = 'HB', repl = 'HB (hard black)', string = statement, count = 1)
print(one_replaced)

# Search and replace using a regex pattern instead of a fixed string
regex_replaced = re.sub(pattern = r'(?<=\.)\s{1}', repl = '\n', string = statement)
print(regex_replaced)
```

#### R {-}

The standard-issue string-search function is `grep()`; it returns the index of the elements in a set of one or more strings for which a pattern match was found. (`grepl()` acts similarly but returns a vector of `TRUE`/`FALSE` indicating whether a match was found in each string passed to the function.) The functions `sub()` and `gsub()` can be used to find and replace instances of a pattern: The former replaces just the first instance; the latter replaces all instances. The search pattern can be provided as a raw character string or as a regular expression.

```{r}
statements <- c('Great Pencil Co. primarily sells pencils of the following grades: HB; B; and 3B.',
                'Great Pencil Co. has its headquarters in Maine, and Great Pencil Co. has supplied the Northeast for decades.')

# Search for pattern and return indexex of elements for which match is found
grep(pattern = 'pencil', x = statements) # When searched for case sensitively, "pencil" is only found in the first string
grep(pattern = '(?i)pencil', x = statements) # When searched for case insensitively, "P/pencil" is found in both strings

# Replace the first instance of a pattern (Co. --> Company)
revised <- sub(pattern = 'Co.', replacement = 'Company', x = statements)
revised

# Replace all instances of a pattern (; --> ,)
revised2 <- gsub(pattern = ';', replacement = ',', x = revised)
revised2

# Find and replace a pattern using regex (3B --> 2B)
final <- sub(pattern = '\\d{1}', replacement = '2', x = revised2)
final
```

Those functions can be used to trim excess (or all) white space in character strings.

```{r}
spaced_string <- c('This      string    started  out with too    many    spaces.')
# Replace all instances of >=2 spaces with single spaces
gsub(pattern = '\\s{2,}', replacement = ' ', x = spaced_string)
# Remove all white space
collapse_these <- c('9:00 - 10:15', '10:15 - 11:30', '11:30 - 12:00')
gsub(pattern = '\\s', replacement = '', x = collapse_these)
```

The package **stringi** also provides an array of string-search and string-manipulation functions, including `stri_detect()`, `stri_replace()`, and `stri_extract()`, all of which easily handle fixed and regex search patterns. For example:

```{r}
library(stringi)
user_dat <- data.frame(name = c('Shire, Jane E', 'Winchester, Marcus L', 'Fox, Sal'), id_number = c('aaa101', 'aaa102', 'aaa103'))
user_dat
# Say we want to use regex patterns and the stringi package to eliminate the 'aaa' patterns from
# the user IDs and then add middle initials---for those users who have them---to the data frame
user_dat$id_number <- stri_replace(user_dat$id_number, regex = '\\w{3}(?=\\d+)', replacement = '')
user_dat$middle_initial <- stri_extract(user_dat$name, regex = '\\b\\w{1}\\b')
user_dat
```

## Change case

How to change the case of strings. The most common case transformations are lower case, upper case, and title case.

#### Python {-}

The `lower()`, `upper()`, and `title()` functions convert case to lower, upper, and title, respectively. We can use a list comprehension to apply these functions to each string in a list.

```{python}
col_names = [col.upper() for col in mtcars.columns]
mtcars.columns = col_names
```

#### R {-}

The `tolower()` and `toupper()` functions convert case to lower and upper, respectively.

```{r}
names(mtcars) <- toupper(names(mtcars))
names(mtcars)
```

```{r}
names(mtcars) <- tolower(names(mtcars))
names(mtcars)
```

The **stringr** package provides a convenient title case conversion function, `str_to_title()`, which capitalizes the first letter of each string.

```{r}
stringr::str_to_title(names(mtcars))
```

## Drop duplicate rows

How to find and drop duplicate elements.

#### Python {-}

The `duplicated()` function determines which rows of a DataFrame are duplicates of previous rows. 

First, we create a DataFrame with a duplicate row by using the pandas `concat()` function. `concat()` combines DataFrames by rows or columns, row by default.

```{python}
# create DataFrame with duplicate rows
import pandas as pd
mtcars2 = pd.concat([mtcars.iloc[0:3,0:6], mtcars.iloc[0:1,0:6]])
```

The `duplicated()` function returns a logical vector. TRUE indicates a row is a duplicate of a previous row.

```{python}
# create DataFrame with duplicate rows
mtcars2.duplicated()
```


#### R {-}

The `duplicated()` function "determines which elements of a vector or data frame are duplicates of elements with smaller subscripts". (from `?duplicated`)

```{r}
# create data frame with duplicate rows
mtcars2 <- rbind(mtcars[1:3,1:6], mtcars[1,1:6])
# last row is duplicate of first
mtcars2
```

The `duplicated()` function returns a logical vector. TRUE indicates a row is a duplicate of a previous row.

```{r}
# last row is duplicate
duplicated(mtcars2)
```

The TRUE/FALSE vector can be used to extract or drop duplicate rows. Since TRUE in indexing brackets will keep a row, we can use `!` to negate the logicals and keep those that are "NOT TRUE"

```{r}
# drop the duplicate and update the data frame
mtcars3 <- mtcars2[!duplicated(mtcars2),]
mtcars3
```

```{r}
# extract and investigate the duplicate row
mtcars2[duplicated(mtcars2),]
```

The `anyDuplicated()` function returns the row number of duplicate rows.

```{r}
anyDuplicated(mtcars2)
```


## Format dates

With formatted dates we can calculate elapsed time, extract components of a date, properly order names of months, and more. 

#### Python {-}
The Python module `datetime` can be used to create various date and time objects. Here we will discuss 4 of the main classes within `datetime` that are most useful. 

The first class we will go over is the `date()` class. This creates a "date" object whose only attributes are year, month, day.

Here we create a date object using the date class. The attributes are specified as integers in the argument of `date()` in this order `date(year, month, day)`. 

```{python}
import datetime as dt 

x = dt.date(2001, 4, 12)
print(x)
```

To get today's date, we can use the `date.today()` function:

```{python}
today = dt.date.today()
print(today)
```

Note that the output of both x and today are only year-month-day because they are date objects. 

We can extract each of these attributes (year, month, day) from the date object as follows: 

```{python}
today.year 
today.month
today.day
```


Next we will discuss the `time()` class. This class creates time objects containing information about only a time. The attributes that go into the `time()` class are hours, minutes, seconds in that order. Like the date class, these attributes must be inputted as integers.  

```{python}
y = dt.time(11, 34, 56)
print(y)
```

If you want a time object containing only hours and minutes, only seconds, etc. you can specify the attributes by name when creating the time object. 

```{python}
only_hrs = dt.time(hour = 10)
only_mins = dt.time(minute = 55)

print(only_hrs)
print(only_mins)
```

Again similar to the date class, we can extract hour, minute, and second attributes from the time objects: 

```{python}
y.hour
y.minute
y.second
y.microsecond
```

Now we will talk about the `datetime()` class that creates a datetime object containing information about both date and time. The attributes must be inputted as integers and are year, month, day, hour, minute, second, in that order. Like the date and time classes, we can speficy specific attributes in the argument using the attrubute names as well. If we don't specify any time components, the datetime object defaults to time 00:00:00. 

```{python}
# Input attributes in order
z = dt.datetime(1981, 4, 12, 11, 34, 56)
print(z)

# Input attributes using attribute names (any order)
z2= dt.datetime(year = 2021, day = 6, month = 12, hour = 6)
print(z2)

# No time attributes
z3 = dt.datetime(1981, 4, 12)
print(z3)
```

Again, we can extract attributes in exactly the same way as the date and time classes. 

```{python}
z.year
z.day
z.hour
```

The final class we will discuss is the `timedelta` class. This class is used to store date/time differences between date objects. 

The default settings for a timedelta object are as follows: `timedelta(weeks=0, days=0, hours=0,minutes=0, seconds=0, milliseconds=0, microseconds=0)`

Here is an example of howto add and subtract datesa and times using these objects.

```{python}
# Create a datetime object for the current time that we will increment
d1 = dt.datetime.now()
print(d1)

# Add 550 days to our datetime object 
d2 = d1 + dt.timedelta(days = 550)
print(d2)

# Subtract 5 hours from our datetime object 
d3 = d1 - dt.timedelta(hours = 5)
print(d3)
```

Finally, we will discuss how to convert strings to datetime objects and vice versa. 

The attribute `strftime()` converts datetime objects to strings. In the argument of `strftime()` can specify the format you would like. 

```{python}
d1

d1.strftime("%A %-m %-Y")
d1.strftime("%a %m %y")
```

The attribute `strptime()` converts strings into datetime objects. In the argument of `strptime()` you must specify the string and then the format of the string. 

```{python}
d4 = "27/10/98 11:03:9.033"

d1.strptime(d4, "%d/%m/%y %H:%M:%S.%f")
```
#### R {-}

Dates in R can be stored as a Date class or a Date-Time class. Dates are stored as the number of days since January 1, 1970. Date-Times are stored as the number of seconds since January 1, 1970. With dates stored in this manner we can calculate elapsed time in units such as days, weeks, hours, minutes, and so forth.

Below are the dates of the first five NASA [Columbia Space Shuttle flights](https://en.wikipedia.org/wiki/List_of_Space_Shuttle_missions#Shuttle_flights) entered as a character vector.

```{r}
date <- c("12 April 1981", 
          "12 November 1981", 
          "22 March 1982", 
          "27 June 1982", 
          "11 November 1982")
```

R does not immediately recognize these as a Date class. To format as a Date class, we can either use the base R `as.Date()` function or one of the convenience functions in the **lubridate** package. The `as.Date()` function requires a specified POSIX conversion specification as documented in `?strptime`. Below the conversion code "%d %B %Y" says Date is entered as two digit day of month (%d), full month name (%B), and year with century (%Y).

```{r}
date1 <- as.Date(date, format = "%d %B %Y")
date1
```

The dates now print in year-month-date format, however they are stored internally as number of days since January 1, 1970. This can be seen by using `as.numeric()` on the "date1" vector.

```{r}
as.numeric(date1)
```

The **lubridate** package provides a series of functions that are permutations of the letters "m", "d", and "y" to represent the order of date components. To format the original "date" vector, we use the `dmy()` function since the date components are ordered as day, month and year. Notice we must load the **lubridate** package to use this function.

```{r message=FALSE}
library(lubridate)
date2 <- dmy(date)
date2
```

When dates are formatted we can easily extract information such as day of week or month. For example to extract the day of week of the launches as an ordered factor, we can use the **lubridate** function `wday()` with `label=TRUE` and `abbr = FALSE`.

```{r}
wday(date2, label = TRUE, abbr = FALSE)
```

To calculate elapsed time between launches in days we can use the base R `diff()` function.

```{r}
diff(date2)
```

To store a date as a Date-Time class we need to include a time component. Below are the first five Columbia launch dates with times. UTC refers to [Universal Coordinated Time](https://en.wikipedia.org/wiki/Coordinated_Universal_Time).

```{r}
datetime <- c("12 April 1981 12:00:04 UTC",
              "12 November 1981 15:10:00 UTC",
              "22 March 1982 16:00:00 UTC",
              "27 June 1982 15:00:00 UTC",
              "11 November 1982 12:19:00 UTC")
```

To format as a Date-Time class we can use either the base R `as.POSIXct()` function or one of the convenience functions in the **lubridate** package. To use `as.POSIXct()` we need to include additional POSIX conversion specifications for the hour, minute and second of launch. The "%H:%M:%S" specification refers to hours, minutes and seconds. The `tz` argument specifies the time zone of the times. 

```{r}
datetime1 <- as.POSIXct(datetime, 
                        format = "%d %B %Y %H:%M:%S", 
                        tz = "UTC")
datetime1
```

When we use `as.numeric()` on the "datetime1" vector we see it is stored as number of seconds since January 1, 1970.

```{r}
as.numeric(datetime1)
```

Using **lubridate** we can append `_hms()` to any of the "mdy" functions to format dates with time components as a Date-Time class. Notice the default time zone in **lubridate** is UTC.

```{r}
datetime2 <- dmy_hms(datetime)
datetime2
```

To calculate elapsed time between launches in hours, we can use the **lubridate** function `time_length()` with the `unit` set to "hours". Below we use `diff()` and then pipe to `time_length()`.

```{r}
diff(datetime2) |> time_length(unit = "hours")
```

For more information on working with dates and times in R, see the vignette accompanying the **lubridate** package. 

## Randomly sample rows

How to take a random sample of rows from a data frame. The sample is usually either a fixed size or a proportion.

#### Python {-}

The pandas package provide a function for taking a sample of fixed size or a proportion. To sample with replacement, set `replace = TRUE`. 

Additionally, the random sample will change every time the code is run. To always generate the same "random" sample, set `random_state` to any positive integer.

To create a sample with a fixed number of rows, use the `n` argument.
```{python}
# sample 5 rows from mtcars
mtcars.sample(n=5, replace=True)
```

To create a sample of a proportion, use the `frac` argument.
```{python}
# sample 20% of rows from mtcars
mtcars.sample(frac = 0.20, random_state=1)
```

The numpy function `random.choice()` in combination with the `loc()` function can be used to sample from a DataFrame. 

The `random.choice()` function creates a random sample according to the given parameters. The `loc()` function is used to access rows and columns by index. 

```{python}
# import the numpy package
import numpy as np

# create a random sample of size 5 with replacement
random_sample = np.random.choice(len(mtcars), (5,), replace=True)

# use random_sample to sample from mtcars
mtcars.loc[random_sample,]
```

The random sample will change every time the code is run. To always generate the same "random" sample, use the `random.seed()` function with any positive integer.

```{python}
# setting seed to always get same random sample
np.random.seed(123)

# create a random sample of size 5 with replacement
sample = np.random.choice(len(mtcars), (5,), replace=True)
mtcars.loc[sample,]
```

#### R {-}

There are many ways to sample rows from a data frame in R. The [**dplyr**](https://dplyr.tidyverse.org/) package provides a convenience function, `slice_sample()`, for taking either a fixed sample size or a proportion.

```{r}
# sample 5 rows from mtcars
dplyr::slice_sample(mtcars, n = 5)

# sample 20% of rows from mtcars
dplyr::slice_sample(mtcars, prop = 0.20)

```

To sample with replacement, set `replace = TRUE`.

The base R functions `sample()` and `runif()` can be combined to sample fixed sizes or approximate proportions.

```{r}
# sample 5 rows from mtcars
# get random row numbers
i <- sample(nrow(mtcars), size = 5)
# use i to select rows
mtcars[i,]
```

```{r}
# sample about 20% of rows from mtcars
# generate random values on range of [0,1]
i <- runif(nrow(mtcars))
# use i < 0.20 logical vector to 
# select rows that correspond to TRUE
mtcars[i < 0.20,]
```

The random sample will change every time the code is run. To always generate the same "random" sample, use the `set.seed()` function with any positive integer.

```{r}
# always get the same random sample
set.seed(123)
i <- runif(nrow(mtcars))
mtcars[i < 0.20,]
```

