# Data Manipulation

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at various strategies for modifying and deriving variables in data.

## Create strings from numbers

You may have data that is numeric but that needs to be treated as a string. For example, census data often codes age ranges as groups. For example:

- 1 = 0-6
- 2 = 7-12 
- 3 = 13-17

And so on. These numbers represent groups, not a count or measure of anything.

#### Python {-}


#### R {-}

The `as.character()` function takes a vector and converts it to string format.

```{r}
# create a vector of numbers
x <- c(2, 2, 3, 4, 2, 4)
x

# convert to string
xc <- as.character(x)
xc

```


The `factor()` function can also be used to convert a numeric vector into a  categorical variable. The result is not exactly a string, however. A factor is made of integers with character labels. Factors are useful for character data that have a fixed set of levels (eg, "grade 1", grade 2", etc)

```{r}
# convert to factor
xf <- factor(x)
xf

# convert to factor with labels
xf2 <- factor(x, labels = c("grade 2", "grade 3", "grade 4"))
xf2

```


## Create numbers from strings


#### Python {-}


#### R {-}


## Change case


#### Python {-}


#### R {-}
