# Data Manipulation

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = '', prompt = FALSE, collapse = FALSE)
```

This chapter discusses strategies for filtering, selecting, modifying, and deriving variables in data. Unless otherwise stated, examples are for DataFrames (Python) and data frames (R) and use the `mtcars` data frame that is included with R.

```{python}
# Python
import pandas
mtcars = pandas.read_csv('data/mtcars.csv')
```


```{r}
# R
data(mtcars)
# Drop row names to match Python version of data
rownames(mtcars) <- NULL
```


## View variable names and types

View and inspect the names of variables and their types (numeric, string, logical, etc.).

#### Python {-}

The `.info()` function in **pandas** returns information about a DataFrame.

Setting the argument `verbose` to `True` prints the names of the columns, their lengths excluding `NULL` values, and their data types (`dtype`) in a table. The function lists the unique data types in the DataFrame, and it prints the amount of memory the DataFrame occupies.

```{python}
mtcars.info(verbose = True)
```

Setting `verbose` to `False` excludes the table describing each column.

```{python}
mtcars.info(verbose = False)
```

If a DataFrame has 100 or fewer columns, the `verbose` argument defaults to `True`. 

#### R {-}

The `str()` function in R lists the names of the variables, their types, the first few values of each, and the object dimensions.

```{r}
str(mtcars)
```

To see just the names of the data frame, use the `names()` function.

```{r}
names(mtcars)
```

To see just the dimensions of the data frame, use the `dim()` function. It returns the number of rows and columns, respectively.


```{r}
dim(mtcars)
```

## Select variables

How to select specific columns of data frames.

#### Python {-}
The period operator `.` provides access to a column in a DataFrame as a vector. This returns a **pandas** Series. A **pandas** Series can do everything a **NumPy** array can do.

```{python}
mtcars.mpg
```

Indexing also provides access to columns as a **pandas** Series. Single and double quotations both work.

```{python}
mtcars['mpg']
```

Operations on **NumPy** arrays are faster than operations on **pandas** Series. But in many cases, using **pandas** Series is fine performance-wise. This consideration is primarily important for large data sets on which many operations are performed. The `.values` function returns a **NumPy** array from a **pandas** DataFrame.

```{python}
mtcars['mpg'].values
```

Double indexing returns a **pandas** DataFrame instead of a **NumPy** array or **pandas** Series.

```{python}
mtcars[['mpg']]
```

The `head()` and `tail()` functions return the first five or last five values, respectively. Use the `n` argument to change the number of values. This function works on **NumPy** arrays, **pandas** Series and **pandas** DataFrames.

```{python}
# First six values
mtcars.mpg.head()
```

```{python}
# Last row of DataFrame
mtcars.tail(n = 1)
```

#### R {-}

The dollar sign operator, `$`, provides access to a column in a data frame as a vector.

```{r}
mtcars$mpg
```

Double-indexing brackets also provide access to columns as a vector.

```{r}
mtcars[["mpg"]]
```

Single-indexing brackets work as well, but they return a data frame instead of a vector (if used with a data frame).

```{r}
mtcars["mpg"]
```


Single-indexing brackets also allow selection of rows when used with a comma. The syntax is `rows, columns`

```{r}
# First three rows
mtcars[1:3, "mpg"]
```

Finally single-indexing brackets allow us to select multiple columns. Request columns either by name or position using a vector.

```{r}
mtcars[c("mpg", "cyl")] 
# Equivalent to mtcars[1:2] 
```

The `head()` and `tail()` functions return the first six or last six values, respectively. Use the `n` argument to change the number of values. These functions work with vectors, data frames, matrices, and more.

```{r}
# First six values
head(mtcars$mpg)
```

```{r}
# Last row of data frame
tail(mtcars, n = 1)
```


## Filter/Subset variables

Selecting rows of a rectangular data set that meet certain conditions.

#### Python {-}

We can filter rows of a DataFrame based on subsetting conditions; the data type returned depends on the filtration method.

The following example returns a DataFrame, not a Series, as there is more than one column selected from the DataFrame. Use a list, square brackets [], to subset more than one column.

```{python}
mtcars[mtcars["mpg"] > 30][["mpg", "cyl"]]
```

Both **pandas** Series and **NumPy** arrays can be used for fast performance and vector operations. Many functions require a vector as input.

The following code returns one column, `mpg`, as a **pandas** Series. A **pandas** Series is one column from a **pandas** DataFrame.

```{python}
mtcars[mtcars["mpg"] > 30]["mpg"]
```

The following code also returns a **pandas** Series, but it uses the `.` operator to select a column, rather than square brackets `[]`.

```{python}
mtcars[mtcars["mpg"] > 30].mpg
```

Both of the following lines of code return **NumPy** arrays using the `.values` function. `df1` is one dimension, reflecting just the `mpg` column, and `df2` is two dimensions, reflecting the `mpg` and `cyl` columns.

```{python}
df1 = mtcars[mtcars["mpg"] > 30]["mpg"].values
df2 = mtcars[mtcars["mpg"] > 30][["mpg", "cyl"]].values
```

You can also filter using multiple row conditions:

```{python}
mtcars[mtcars["mpg"] > 30][mtcars["hp"] < 66]
```

#### R {-}

In base R, we can use subsetting brackets or the `subset()` function to select rows based on some condition. Below, we demonstrate both approaches, returning only those rows with `mpg` greater than 30. We begin with subsetting brackets, which take three arguments: 

1. `i`: the condition to subset on.
2. `j`: the columns to show. If none specified, all columns are returned
3. `drop`: an optional logical argument (TRUE/FALSE) to determine whether or not to coerce the output to the lowest possible dimension. The default is TRUE.

We rarely type the first two argument names, `i` and `j`, when using subsetting brackets.

This example returns only the rows with mpg > 30 and all columns. Notice we need to preface mpg with `mtcars$` to tell R where to find the "mpg" column and that we need to provide a comma after the condition.

```{r}
mtcars[mtcars$mpg > 30, ]
```

We can select the columns to return in the second argument as a vector. Notice that we only need to specify the column names as a character vector. We can also use numbers corresponding to the column positions, as well as conditional statements.

```{r}
mtcars[mtcars$mpg > 30, c("mpg", "wt", "gear")]
```

Show the first three columns:

```{r}
mtcars[mtcars$mpg > 30, 1:3]
```

Show columns with names consisting of only two characters. The `nchar()` function counts the number of characters in a string. The expression `nchar(names(mtcars)) == 2` returns a vector of `TRUE`/`FALSE` values where `TRUE` indicates the column name is only two characters in length.

```{r}
mtcars[mtcars$mpg > 30, nchar(names(mtcars)) == 2]
```

Notice that when we specify only one column, the brackets return a vector.

```{r}
mtcars[mtcars$mpg > 30, "mpg"]
```

To get a data frame, set the `drop` argument to FALSE.

```{r}
mtcars[mtcars$mpg > 30, "mpg", drop = FALSE]
```

The `subset()` function allows us to refer to column names without using the `$` extractor function or quoting column names. It also has a drop argument, but its default is `FALSE.` It has four arguments:

1. `x`: the data frame to subset.
2. `subset`: the condition to subset on.
3. `select`: the columns to select.
4. `drop`: an optional logical argument (TRUE/FALSE) to determine whether or not to coerce the output to the lowest possible dimension. The default is FALSE.

We rarely type the first three argument names, `x`, `subset` and `select`, when using `subset()`.

Below we replicate the previous examples using `subset()`.

```{r}
# Return rows where mpg > 30 and all columns
subset(mtcars, mpg > 30)
```

```{r}
# Return rows where mpg > 30 and the mpg, wt, and gear columns
subset(mtcars, mpg > 30, c(mpg, wt, gear))
```

```{r}
# Return rows where mpg > 30 and the first three columns
subset(mtcars, mpg > 30, 1:3)
```

```{r}
# Return rows where mpg > 30 and columns consisting of two characters
subset(mtcars, mpg > 30, nchar(names(mtcars)) == 2)
```

```{r}
# Return rows where mpg > 30 and mpg column, as a vector
subset(mtcars, mpg > 30, mpg, drop = TRUE)
```

```{r}
# Return rows where mpg > 30 and mpg column, as a data frame
subset(mtcars, mpg > 30, mpg)
```

Another difference between subsetting brackets and the `subset()` function is how they handle missing values. Subsetting brackets return missing values while `subset()` does not. We demonstrate below with a toy data frame. Notice the `x` column has a missing value.

```{r}
dframe <- data.frame(x = c(1, NA, 5), 
                     y = c(12, 21, 34))
dframe
```

When we condition on `x < 3`, the subsetting bracket approach returns a row with `NA` values.

```{r}
dframe[dframe$x < 3, ]
```

The `subset()` approach ignores the missing value.

```{r}
subset(dframe, x < 3)
```

To replicate the `subset()` result with the subsetting brackets, we need to include an additional condition to only show rows where x is _not_ missing. We can do that with the `is.na()` function. The `is.na()` function returns `TRUE` if a value is missing and `FALSE` otherwise. If we preface with `!`, we get `TRUE` if a value is _not_ missing and `FALSE` otherwise. 

```{r}
dframe[dframe$x < 3 & !is.na(dframe$x),]
```

See also the `filter()` function in the [**dplyr**](https://dplyr.tidyverse.org/) package and the enhanced subsetting brackets in the [**data.table**](https://rdatatable.gitlab.io/data.table/) package.

## Rename variables

#### Python {-}

Column names can be changed using the function `.rename()`. Below, we change the column names `cyl` and `wt` to `cylinder` and `WT`, respectively.

```{python}
mtcars.rename(columns = {"cyl":"cylinder", "wt":"WT"})
```

Alternatively, column names can be changed by replacing the vector of column names with a new vector. Below, we create a vector of columns that replaces `drat` with `axle_ratio` using conditional match and that replaces `disp` with `DISP` using indexing.

```{python}
column_names = mtcars.columns.values

# Using conditional match
column_names[column_names == "drat"] = "axle_ratio"

# Using indexing
column_names[2] = "DISP"

mtcars.columns = column_names
mtcars.columns
```

#### R {-}

Variable names can be changed by referencing their indexes (i.e., their positions among the columns in a data frame). Below the second column is `cyl`. We change the name to `cylinders.`

```{r}
names(mtcars)[2]
names(mtcars)[2] <- "cylinders"
names(mtcars)
```

Variable names can also be changed by conditional match. Below we find the variable name that matches `drat` and change it to `axle_ratio.`

```{r}
names(mtcars)[names(mtcars) == "drat"]
names(mtcars)[names(mtcars) == "drat"] <- "axle_ratio"
names(mtcars)
```

More than one variable name can be changed using a vector of positions or matches.

```{r}
names(mtcars)[c(6,8)] <- c("weight", "engine")
# Alternatively: names(mtcars)[names(mtcars) %in% c("wt", "vs")] <- c("weight", "engine")

names(mtcars)
```

See also the `rename()` function in the [**dplyr**](https://dplyr.tidyverse.org/) package.

## Create, replace and remove variables

We often need to create variables that are functions of other variables, or we need to replace existing variables with updated versions thereof.

#### Python {-}

Adding a new variable via indexing notation and assigning a result to it adds a new column.

```{python}
# Add column for kilometer per liter
mtcars['kpl'] = mtcars.mpg / 2.352
```

Doing the same with an _existing_ column updates the values in the column.

```{python}
# Update to liters per 100 kilometers
mtcars['kpl'] = 100 / mtcars.kpl 
```

Alternatively, the `.` notation can be used to update the values in a column.
```{python}
# Update to liters per 50 Kilometers
mtcars.kpl = 50 / mtcars.kpl 
```

To remove a column, use the `.drop()` function.

```{python}
# Drop the kpl variable
mtcars.drop(columns = ['kpl'])
```

#### R {-}

Adding a new variable name after the dollar sign notation and assigning a result to it adds a new column.

```{r}
# Add column for kilometer per liter
mtcars$kpl <- mtcars$mpg / 2.352
```

Doing the same with an _existing_ variable updates the values in a column. 

```{r}
# Update to liters per 100 Kilometers
mtcars$kpl <- 100 / mtcars$kpl 
```

To remove a variable, assign it `NULL`.

```{r}
# Drop the kpl variable
mtcars$kpl <- NULL
```

See also the `mutate()` function in the [**dplyr**](https://dplyr.tidyverse.org/) package.

## Create strings from numbers

You may have data that is numeric but that needs to be treated as a string. 

#### Python {-}

You can change the data type of a column in a DataFrame using the `astype` function.

```{python}
mtcars['am'] = mtcars['am'].astype(str)
type(mtcars.am[0]) # check the type of the first item in am column
```

A potential number-to-string conversion task in Python might be formatting 5-digit American ZIP Codes. Some zip codes begin with 0, but if stored as a numeric value, the 0 is dropped. For example, consider the following **pandas** DataFrame. Notice that the leading 0 is dropped from two of the ZIP Codes.

```{python}
zc = pandas.read_csv('data/zc.csv')
print(zc)
```

One way to fix this is using the string `zfill()` method. First, we convert the numeric column to string type using the method we just demonstrated. Then we access the `zip` column using `zc.zip` and the `zfill()` method using `str.zfill` with the width parameter set to 5. This pads the string with `0` on the left as necessary to make each value five characters wide. 

```{python}
zc['zip'] = zc['zip'].astype(str)
zc['zip'] = zc.zip.str.zfill(5)
print(zc)
```

If we knew we were importing zip codes using `read_csv`, we could also use the `dtype` argument to specify which storage type to use for the `zip` column. Below, we pass a dictionary that maps the `str` type to the `zip` column. The result is a properly formatted ZIP Code column.

```{python}
zc = pandas.read_csv('data/zc.csv', dtype = {'zip':'str'})
print(zc)
```

#### R {-}

The `as.character()` function takes a vector and converts it to string format.

```{r}
head(mtcars$am)
head(as.character(mtcars$am))
```

To save the results of this process---this conversion---we need to _assign_ the result to the data frame.

```{r}
# Add new string variable am_ch
mtcars$am_ch <- as.character(mtcars$am)
head(mtcars$am_ch)
```

The `factor()` function can also be used to convert a numeric vector into a  categorical variable. However, the result is not exactly a string: A factor is made of integers with character labels. Factors are useful for character data that have a fixed set of levels (e.g., "grade 1," grade 2," etc.).

```{r}
# Convert to factor
head(mtcars$am)
head(factor(mtcars$am))

# Convert to factor with labels
head(factor(mtcars$am, labels = c("automatic", "manual")))
```

As before, to save the results of this conversion to the data frame, we need to assign the results to a variable:

```{r}
# Create factor variable am_fac
mtcars$am_fac <- factor(mtcars$am, labels = c("automatic", "manual"))
head(mtcars$am_fac)
```

A common number-to-string conversion task in R is formatting 5-digit American ZIP Codes as demonstrated above in Python. Some zip codes begin with 0, but if stored as a numeric value, the 0 is dropped.

```{r}
zip_codes <- c(03766, 03748, 22901, 03264)
zip_codes
```

We need to store the zip code as a character value so the 0 is preserved. One way to do this is via the `sprintf()` function in base R. The first argument is the _format string_ or _conversion specification_. A conversion specification begins with `%`. The following `0` and `5` says to format the zip_codes vector as a 5-digit string padded by zeroes on the left. The final `i` says we're working with integer values. 

```{r}
sprintf("%05i", zip_codes)
```

See also the `str_pad()` function in the **stringr** package.

## Create numbers from strings

String variables that ought to be numbers usually have some character data contained in the values, such as units (e.g., "4 cm"). To create numbers from strings, it's important to remove any character data that cannot be converted to a number.

#### Python {-}

The `astype(float)` or `astype(int)` function will coerce strings to numerical representation.

For demonstration, let's say we have the following **NumPy** array:

```{python}
import numpy as np
weight = np.array(["125 lbs.", "132 lbs.", "156 lbs."])
```

The `astype(float)` function throws an error due to the presence of strings. The `astype()` function is for **NumPy** arrays.

```{python}
try:
  weight.astype(float)
except ValueError:
  print("ValueError: could not convert string to float: '125 lbs.'")
```

One way to approach this is to first remove the strings from the objects and then use `astype(float)`. Below we use the `strip()` function to find " lbs." using a list comprehension.

```{python}
# [] indicates a list in python
# np.array() changes the list back into an array
weight = np.array([w.strip(" lbs.") for w in weight])
```

Now we can use the `astype()` function to change the elements in weight from `str` to `float`.

```{python}
weight.astype(float)
```

#### R {-}

The `as.numeric()` function will attempt to coerce strings to numeric type _if possible_. Any non-numeric values are coerced to `NA`.

For demonstration, say we have the following vector:

```{r}
weight <- c("125 lbs.", "132 lbs.", "156 lbs.")
```

The `as.numeric()` function returns all NA due to presence of character data.

```{r}
as.numeric(weight)
```

There are many ways to approach this. A common approach is to first remove the characters and then use `as.numeric()`. Below, we use the `gsub()` function to find "lbs." and replace it with nothing (find-and-replace procedures are discussed more below).

```{r}
weightN <- gsub(" lbs.", "", weight)
as.numeric(weightN)
```

The `parse_number()` function in the **readr** package can often take care of these situations automatically.

```{r}
readr::parse_number(weight)
```

## Combine strings

String concatenation---turning "Jane" and "Smith" into "Jane Smith"---is easily done in both languages.

#### Python {-}

The `+` operator can combine strings in Python.

```{python}
species = 'yellow-bellied sea snake'
tail_shape = 'paddle-shaped'

statement = 'The ' + species + ' has a ' + tail_shape + ' tail that helps it swim.'
print(statement)
```

#### R {-}

The `paste()` and `paste0()` functions combine strings in R. The former concatenates strings and places spaces between them; the latter concatenates sans spaces.

```{r}
species <- 'rainbow boa'
appearance <- 'iridescent'
location <- 'Central and South America'

statement1 <- paste('The', species, 'has an', appearance, 'sheen.')
statement1

# Note that spaces must be provided explicitly when using paste0()
statement2 <- paste0('The ', species, ' is found in ', location)
statement2
```

## Finding and replacing patterns within strings

This section reviews key functions in Python and R for finding and replacing character patterns. The functions we discuss can search for fixed character patterns (e.g., "Meredith Rollins" to case-sensitively match that name and that name alone) or regular expression (regex) patterns (e.g., `\w+` to capture all instances of >=1 alphabetic character). Note that in R, meta characters like `w` (to match word characters) and `d` (to match digits) are escaped with _two_ backslashes (e.g., `\\w` and `\\d`). In Python, regex patterns are generally headed by `r`, which allows meta characters in the regex itself to be escaped with just one `\` (e.g., `r"\w+"`). Regex is an enormous topic, and we don't discuss it at any length here, but you can learn more about regular expressions---and how they're implemented in different programming languages---at these websites: https://www.regular-expressions.info/; https://regexone.com/

#### Python {-}

The **re** module provides a set of functions for searching and manipulating strings. The `search()` function does exactly as its name suggests: It identifies matches for a fixed or regex character pattern in a string. `sub()` searches for and replaces character patterns (fixed or regex). The `count` argument in `sub()` allows a user to specify how many instances of the matched pattern they want to to replace; e.g., use `count = 1` to replace just the first instance of a match.

```{python}
import re
statement = 'Pencils with an HB graphite grade are commonly used for writing. An HB pencil is approximately equal to a #2 pencil.'

# Search for "HB" using fixed and regex patterns
search_result1 = re.search(pattern = "HB", string = statement)
print(search_result1)
search_result2 = re.search(pattern = r"[H,B]{2}", string = statement)
print(search_result2)
```
```{python}
# Replace all instances of "HB"
all_replaced = re.sub(pattern = 'HB', repl = 'HB (hard black)', string = statement)
print(all_replaced)
```
```{python}
# Replace just the first instance of HB
one_replaced = re.sub(pattern = 'HB', repl = 'HB (hard black)', string = statement, count = 1)
print(one_replaced)
```
```{python}
# Search and replace using a regex pattern instead of a fixed string
# (Replace a space after a period with a line break)
regex_replaced = re.sub(pattern = r'(?<=\.)\s{1}', repl = '\n', string = statement)
print(regex_replaced)
```

#### R {-}

The standard-issue string-search function in base R is `grep()`; it returns the index of the elements in a set of one or more strings for which a pattern match was found. (`grepl()` acts similarly but returns a vector of `TRUE`/`FALSE` indicating whether a match was found in each string passed to the function.) The functions `sub()` and `gsub()` can be used to find and replace instances of a pattern: The former replaces just the first instance; the latter replaces all instances. The search pattern can be provided as a raw character string or as a regular expression.

```{r}
statements <- c('Great Pencil Co. primarily sells pencils of the following grades: HB; B; and 3B.',
                'Great Pencil Co. has its headquarters in Maine, and Great Pencil Co. has supplied the Northeast for decades.')

# Search for a pattern and return the indexes of elements for which a match is found
grep(pattern = 'pencil', x = statements, fixed = TRUE) # use `fixed = T` to search for exactly the declared string: "pencil" in all lowercase, which is only found in one string
grep(pattern = '(?i)pencil', x = statements) # using regex allows for flexible searches, such as a case-insensitive search for "P/pencil", which, unlike "pencil" is found in in both strings

# Replace the first instance of a pattern (Co. --> Company)
revised <- sub(pattern = 'Co\\.', replacement = 'Company', x = statements)
revised

# Replace all instances of a pattern (; --> ,)
revised2 <- gsub(pattern = ';', replacement = ',', x = revised)
revised2

# Find and replace a pattern using regex (3B --> 2B)
final <- sub(pattern = '\\d{1}', replacement = '2', x = revised2)
final
```

Those functions can be used to trim excess (or all) white space in character strings.

```{r}
spaced_string <- c('This      string    started  out with too    many    spaces.')
# Replace all instances of >=2 spaces with single spaces
gsub(pattern = '\\s{2,}', replacement = ' ', x = spaced_string)
# Remove all white space
collapse_these <- c('9:00 - 10:15', '10:15 - 11:30', '11:30 - 12:00')
gsub(pattern = '\\s', replacement = '', x = collapse_these)
```

The package **stringi** also provides an array of string-search and string-manipulation functions, including `stri_detect()`, `stri_replace()`, and `stri_extract()`, all of which easily handle fixed and regex search patterns. For example:

```{r}
library(stringi)
user_dat <- data.frame(name = c('Shire, Jane E', 'Winchester, Marcus L', 'Fox, Sal'),
                       id_number = c('aaa101', 'aaa102', 'aaa103'))
user_dat

# Say we want to to eliminate the 'aaa' patterns from the user IDs and then add
# middle initials---for those users who have them---to the data frame
user_dat$id_number <- stri_replace(user_dat$id_number, regex = '^\\w{3}(?=\\d+)', replacement = '') # alternatively: stri_extract(user_dat$id_number, regex = '(?<=\\w{3})\\d+$')

user_dat$middle_initial <- stri_extract(user_dat$name, regex = '\\b\\w{1}$')
user_dat
```

## Change case

How to change the case of strings. The most common case transformations are lower case, upper case, and title case.

#### Python {-}

The `lower()`, `upper()`, and `title()` functions convert strings to lower, upper, and title case, respectively. We can use a list comprehension to apply these functions to each string in a list.

```{python}
col_names = [col.upper() for col in mtcars.columns]
mtcars.columns = col_names
```

#### R {-}

The `tolower()` and `toupper()` functions convert case to lower and upper, respectively.

```{r}
names(mtcars) <- toupper(names(mtcars))
names(mtcars)
```

```{r}
names(mtcars) <- tolower(names(mtcars))
names(mtcars)
```

The **stringr** package provides a convenient title-case conversion function, `str_to_title()`, which capitalizes the first letter of each string.

```{r}
stringr::str_to_title(names(mtcars))
```

## Drop duplicate rows

How to find and drop duplicate elements.

#### Python {-}

The `duplicated()` function determines which rows of a DataFrame are duplicates of previous rows. 

First, we create a DataFrame with a duplicate row by using the **pandas** `concat()` function. `concat()` combines DataFrames by rows or columns (by rows is the default).

```{python}
# Create DataFrame with duplicate rows
import pandas as pd
mtcars2 = pd.concat([mtcars.iloc[0:3,0:6], mtcars.iloc[0:1,0:6]])
```

The `duplicated()` function returns a logical vector. TRUE indicates a row is a duplicate of a previous row.

```{python}
mtcars2.duplicated()
```

#### R {-}

The `duplicated()` function "determines which elements of a vector or data frame are duplicates of elements with smaller subscripts" (from `?duplicated`).

```{r}
# Create data frame with duplicate rows
mtcars2 <- rbind(mtcars[1:3,1:6], mtcars[1,1:6])
# Last row is duplicate of first
mtcars2
```

The `duplicated()` function returns a logical vector. `TRUE` indicates that a given element is a duplicate of a previous one (and here, element = row).

```{r}
# The last row is a duplicate
duplicated(mtcars2)
```

The `TRUE`/`FALSE` vector can be used to extract or drop duplicate rows. Since `TRUE` in indexing brackets will keep a row, we can use `!` to negate the logicals and keep those that are _not_ `TRUE.`

```{r}
# Drop the duplicate and update the data frame
mtcars3 <- mtcars2[!duplicated(mtcars2), ]
mtcars3
```

```{r}
# Extract and investigate the duplicate row
mtcars2[duplicated(mtcars2), ]
```

The `anyDuplicated()` function returns the row number of duplicate rows.

```{r}
anyDuplicated(mtcars2)
```

## Format dates

With formatted dates, we can calculate elapsed times, extract components of a date, properly order the names of months, and more. 

#### Python {-}
The Python module **datetime** can be used to create various date and time objects. Here, we discuss four of the main classes within **datetime** 

The first class is the `date()` class. This creates a date object whose only attributes are year, month, and day.

Below, we create a date object using the `date()` class. The attributes are specified as integers in the arguments of `date()` in the following order: `date(year, month, day)`. 

```{python}
import datetime as dt 

x = dt.date(2001, 4, 12)
print(x)
```

To get today's date, we can use the `date.today()` function:

```{python}
today = dt.date.today()
print(today)
```

Note that the output of both `x` and `today` are only year-month-day because they are date objects. 

We can extract each of these attributes (year, month, day) from the date object as follows: 

```{python}
today.year 
today.month
today.day
```

Next is the `time()` class. This class creates time objects containing information about only a time. The attributes that go into the `time()` class are hours, minutes, and seconds in that order. Like the date class, these attributes must be input as integers.  

```{python}
y = dt.time(11, 34, 56)
print(y)
```

If you want a time object containing only hours and minutes, or only seconds, etc. you can specify the attributes by name when creating the time object. 

```{python}
only_hrs = dt.time(hour = 10)
only_mins = dt.time(minute = 55)

print(only_hrs)
print(only_mins)
```

Again, similar to the date class, we can extract hour, minute, and second attributes from time-classed objects: 

```{python}
y.hour
y.minute
y.second
y.microsecond
```

The `datetime()` class that creates a datetime object containing information about both date and time. The attributes must be input as integers wiht the following default order: year, month, day, hour, minute, and second. Like the date and time classes, we can specify specific attributes in the arguments using the names thereof. If we don't specify any time components, the datetime object defaults to time 00:00:00. 

```{python}
# Input attributes in order
z = dt.datetime(1981, 4, 12, 11, 34, 56)
print(z)
```
```{python}
# Input attributes using attribute names (any order)
z2= dt.datetime(year = 2021, day = 6, month = 12, hour = 6)
print(z2)
```
```{python}
# No time attributes
z3 = dt.datetime(1981, 4, 12)
print(z3)
```

Again, we can extract attributes in exactly the same way as we did for the date and time classes. 

```{python}
z.year
z.day
z.hour
```

The final class discussed here is the timedelta class. This class is used to store date/time differences between date objects. 

The default settings for a timedelta object are as follows: `timedelta(weeks = 0, days = 0, hours = 0, minutes = 0, seconds = 0, milliseconds = 0, microseconds = 0)`.

To add or subtract dates and times to/from these objects, we can use operators like `+` and `-`:

```{python}
# Create a datetime object for the current time
d1 = dt.datetime.now()
print(d1)
```
```{python}
# Add 550 days to the datetime object 
d2 = d1 + dt.timedelta(days = 550)
print(d2)
```
```{python}
# Subtract five hours from the datetime object 
d3 = d1 - dt.timedelta(hours = 5)
print(d3)
```

Finally, we will discuss how to convert strings to datetime objects and vice versa. 

The attribute `strftime()` converts datetime objects to strings. In the argument of `strftime()`, you can specify the format you would like. 

```{python}
d1

d1.strftime("%A %m %Y")
d1.strftime("%a %m %y")
```

The attribute `strptime()` converts strings into datetime objects. In the argument of `strptime()`, you must specify the string and then the format of the string. 

```{python}
d4 = "27/10/98 11:03:9.033"

d1.strptime(d4, "%d/%m/%y %H:%M:%S.%f")
```
#### R {-}

Dates in R can be stored as a Date class or a Date-Time class. Dates are stored as the number of days since January 1, 1970. Date-Times are stored as the number of seconds since January 1, 1970. With dates stored in this manner, we can calculate elapsed time in units such as days, weeks, hours, minutes, and so forth.

Below are the dates of the first five NASA [Columbia Space Shuttle flights](https://en.wikipedia.org/wiki/List_of_Space_Shuttle_missions#Shuttle_flights) entered as a character vector.

```{r}
date <- c("12 April 1981", 
          "12 November 1981", 
          "22 March 1982", 
          "27 June 1982", 
          "11 November 1982")
```

R does not immediately recognize these as a Date class. To format as a Date class, we can use either the base R `as.Date()` function or one of the convenience functions in the **lubridate** package. The `as.Date()` function requires a specified POSIX conversion specification as documented in `?strptime`. Below, the conversion code `%d %B %Y` indicates that the date is entered as a two-digit day of month (`%d`), a full month name (`%B`), and a year with century (`%Y`).

```{r}
date1 <- as.Date(date, format = "%d %B %Y")
date1
```

The dates now print in year-month-day format; however, they are stored internally as number of days since January 1, 1970. This can be seen by using `as.numeric()` on the `date1` vector.

```{r}
as.numeric(date1)
```

The **lubridate** package provides a series of functions that are permutations of the letters "m", "d", and "y" to represent the order of date components. To format the original "date" vector, we use the `dmy()` function since the date components are ordered as day, month and year. We must load the **lubridate** package to use this function.

```{r message=FALSE}
library(lubridate)
date2 <- dmy(date)
date2
```

When dates are formatted, we can easily extract information such as the day of week or the month. For example, to extract the day of week of the launches as an ordered factor, we can use the **lubridate** function `wday()` with `label = TRUE` and `abbr = FALSE`.

```{r}
wday(date2, label = TRUE, abbr = FALSE)
```

To calculate elapsed time between launches in days, we can use the base R `diff()` function.

```{r}
diff(date2)
```

To store a date as a Date-Time class, we need to include a time component. Below are the first five Columbia launch dates with times. (UTC refers to [Universal Coordinated Time](https://en.wikipedia.org/wiki/Coordinated_Universal_Time).)

```{r}
datetime <- c("12 April 1981 12:00:04 UTC",
              "12 November 1981 15:10:00 UTC",
              "22 March 1982 16:00:00 UTC",
              "27 June 1982 15:00:00 UTC",
              "11 November 1982 12:19:00 UTC")
```

To format as a Date-Time class, we can use either the base R `as.POSIXct()` function or one of the convenience functions in the **lubridate** package. To use `as.POSIXct()`, we need to include additional POSIX conversion specifications for the hour, minute, and second of launch. The `%H:%M:%S` specification refers to hours, minutes, and seconds. The `tz` argument specifies the time zone of the times. 

```{r}
datetime1 <- as.POSIXct(datetime, 
                        format = "%d %B %Y %H:%M:%S", 
                        tz = "UTC")
datetime1
```

When we use `as.numeric()` on the `datetime1` vector, we see it is stored as number of seconds since January 1, 1970.

```{r}
as.numeric(datetime1)
```

Using **lubridate**, we can append `_hms()` to any of the "mdy" functions to format dates with time components as a Date-Time class. Notice the default time zone in **lubridate** is UTC.

```{r}
datetime2 <- dmy_hms(datetime)
datetime2
```

To calculate the elapsed time between launches in hours, we can use the **lubridate** function `time_length()` with `unit = "hours"`. Below, we use `diff()` and then pipe the result to `time_length()`.

```{r}
diff(datetime2) |> time_length(unit = "hours")
```

For more information on working with dates and times in R, see the vignette accompanying the **lubridate** package. 

## Randomly sample rows

How to take a random sample of rows from a data frame. The sample is usually either a fixed size or a proportion.

#### Python {-}

The **pandas** package provides a function for taking a sample of fixed size or a proportion, `sample()`. To sample with replacement, set `replace = TRUE`. 

By default, the random sample will change every time the code is run. To always generate the same "random" sample, set `random_state` to any positive integer.

To create a sample with a fixed number of rows, use the `n` argument.
```{python}
# Sample five rows from mtcars
mtcars.sample(n = 5, replace = True)
```

To create a sample of a proportion, use the `frac` argument.
```{python}
# Sample 20% of the rows from mtcars
mtcars.sample(frac = 0.20, random_state = 1)
```

The **NumPy** function `random.choice()` in combination with the `loc()` function can be used to sample from a DataFrame. 

The `random.choice()` function creates a random sample according to the given parameters. The `loc()` function is used to access rows and columns by index. 

```{python}
import numpy as np

# Create a random sample of size five with replacement
random_sample = np.random.choice(len(mtcars), (5, ), replace=True)

# Use random_sample to sample from mtcars
mtcars.loc[random_sample, ]
```

As before, setting `random.seed()` to a positive integer will ensure that the same "random" sample is generated each time the code is run.

```{python}
np.random.seed(123)
sample = np.random.choice(len(mtcars), (5,), replace=True)
mtcars.loc[sample, ]
```

#### R {-}

There are many ways to sample rows from a data frame in R. The [**dplyr**](https://dplyr.tidyverse.org/) package provides a convenience function, `slice_sample()`, for taking either a fixed sample size or a proportion.

```{r}
# Sample five rows from mtcars
dplyr::slice_sample(mtcars, n = 5)

# Sample 20% of rows from mtcars
dplyr::slice_sample(mtcars, prop = 0.20)

```

To sample with replacement, set `replace = TRUE`.

The base R functions `sample()` and `runif()` can be combined to sample fixed sizes or approximate proportions.

```{r}
# Sample five rows from mtcars
# First, get five random row numbers
i <- sample(nrow(mtcars), size = 5)
# Then, use i to subset those rows
mtcars[i, ]
```

```{r}
# Sample about 20% of rows from mtcars
# First, generate random values on range of [0,1]
i <- runif(nrow(mtcars))
# Then, use i < 0.20 logical vector to select rows that correspond to TRUE
mtcars[i < 0.20, ]
```

The random sample will change every time the code is run. To always generate the same "random" sample, use the `set.seed()` function with any positive integer.

```{r}
set.seed(123)
i <- runif(nrow(mtcars))
mtcars[i < 0.20,]
```