# Combine, Reshape and Merge

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at various strategies for combining, reshaping, and merging data.

## Combine rows

Combining rows may be thought of as "stacking" rectangular data structures.

#### Python {-}


#### R {-}

The `rbind()` function "binds" rows. It takes two or more objects. To row bind data frames the column names must match, otherwise an error is returned. If  columns being stacked have differing variable types, the values will be coerced according to `logical` < `integer` < `double` < `complex` < `character`. (E.g., if you stack a set of rows with type `logical` in column _J_ on a set of rows with type `character` in column _J_, the output will have column _J_ as type `character`.)

```{r}
d1 <- data.frame(x = 4:6, y = letters[1:3])
d2 <- data.frame(x = 3:1, y = letters[4:6])
rbind(d1, d2)
```

See also the `bind_rows()` function in the **dplyr** package.

## Combine columns

Combining columns may be thought of as setting rectangular data structures next to each other.

#### Python {-}


#### R {-}

The `cbind()` function "binds" columns. It takes two or more objects. To column bind data frames, the number of rows must match; otherwise, the object with fewer rows will have rows "recycled" (if possible) or an error will be returned.

```{r}
d1 <- data.frame(x = 10:13, y = letters[1:4])
d2 <- data.frame(x = c(23,34,45,44))
cbind(d1, d2)
```

```{r}
# example of recycled rows (d1 is repeated twice)
d1 <- data.frame(x = 10:13, y = letters[1:4])
d2 <- data.frame(x = c(23,34,45,44,99,99,99,99))
cbind(d1, d2)
```

See also the `bind_cols()` function in the **dplyr** package.

## Reshaping data

The next two sections discuss how to reshape data from wide to long and from long to wide. "Wide" data are structured such that multiple values associated with a given unit (e.g., a person, a cell culture, etc.) are placed in the same row:
  
```{r, echo = F}
temp <- data.frame(name = c('larry', 'moe', 'curly'), time_1_score = c(3, 6, 2), time_2_score = c(0, 3, 1))
temp
```

_Long_ data, conversely, are structured such that all values are contained in one column, with another column identifying what value is given in any particular row ("time 1," "time 2," etc.):

```{r, echo = F}
temp <- data.frame(id = rep(c('larry', 'moe', 'curly'), each = 2), time = rep(1:2, times = 3), score = c(3, 0, 6, 3, 2, 1))
temp
```

Shifting between these two data formats is often necessary for implementing certain statistical techniques or representing data with particular visualizations.

### Wide to long

#### Python {-}

#### R {-}

In base R, the `reshape()` function can take data from wide to long or long to wide. The **tidyverse** also provides reshaping functions: `pivot_longer()` and `pivot_wider()`. The **tidyverse** functions have a degree of intuitiveness and usability that may make them the go-to reshaping tools for many R users. We give examples below using both base R and **tidyverse**.

Say we begin with a wide data frame, `df_wide`, that looks like this:
```{r, echo = F}
set.seed(10)
df_wide <- data.frame(id = rep(1:3),
                      sex = sample(c('m', 'f'), 3, replace = T),
                      wk1 = sample(1:20, 3, replace = T),
                      wk2 = sample(1:20, 3, replace = T),
                      wk3 = sample(1:20, 3, replace = T))
df_wide
```

To lengthen a data frame using `reshape()`, a user provides arguments specifying the columns that identify values' origins (person, cell culture, etc.), the columns containing values to be lengthened, and the desired names for output columns in long data:

```{r}
df_long <- reshape(df_wide,
                        direction = 'long',
                        idvar = c('id', 'sex'), # column(s) that uniquely identifies/y each row
                        varying = c('wk1', 'wk2', 'wk3'), # variables that contain the values to be lengthened
                        v.names = 'val', # desired name of column in long data that will contain values
                        timevar = 'week') # desired name of column in long data that will identify each value's context
df_long
```

The **tidyverse** function for taking data from wide to long is `pivot_longer()`. To lengthen `df_wide` using `pivot_longer()`, a user would write:
  
```{r, message = F}
library(tidyverse)
df_long_PL <- pivot_longer(df_wide,
                             cols = -c('id', 'sex'), # columns that contain the values to be lengthened (can use -c() to negate variables)
                             names_to = 'week', # desired name of column in long data that will identify each value's context
                             values_to = 'val') # desired name of column in long data that will contain values
df_long_PL
```

`pivot_longer()` is particularly useful (a) when dealing with wide data that contain multiple sets of repeated measures in each row that need to be lengthened separately (e.g., two monthly height measurements and two monthly weight measurements for each person) and (b) when column names and/or column values in the long data need to be extracted from column names of the wide data using regular expressions.

For example, say we begin with a wide data frame, `animals_wide`, in which every row contains two values for each of two different measures:

```{r, echo = F}
library(tidyverse)
animals_wide <- data.frame(animal = c('dolphin', 'porcupine', 'capybara'),
                           lives_in_water = c(T, F, F),
                           jan_playfulness = c(6, 3.5, 4.0),
                           feb_playfulness = c(5.5, 4.5, 5.0),
                           jan_excitement = c(7.0, 3.5, 4.0),
                           feb_excitement = c(7.0, 3.5, 4.0))
animals_wide
```

`pivot_longer()` can be used to convert this data frame to a long format where there is one column for each of the measures, playfulness and excitement:

```{r}
animals_long_1 <- pivot_longer(animals_wide,
                             cols = -c('animal', 'lives_in_water'),
                             names_to = c('.value', 'measure'), # ".value" is placeholder for values that will be extracted from wide column names 
                             names_pattern = '(.+)_(.+)') # specify structure of wide column names with regex from which long column names will be extracted
animals_long_1
```

Alternatively, `pivot_longer()` can be used to convert this data frame to a long format where there is one column containing all the playfulness and excitement values:

```{r}
animals_long_2 <- pivot_longer(animals_wide,
                               cols = -c('animal', 'lives_in_water'),
                               names_to = c('month', 'measure'),
                               names_pattern = '(.+)_(.+)',
                               values_to = 'val')
animals_long_2
```

### Long to wide

#### Python {-}

#### R {-}

Say we begin with a long data frame, `df_long`, that looks like this:

```{r}
df_long
```

To take data from long to wide with base R's `reshape()`, a user would write:

```{r}
df_wide <- reshape(df_long,
                   direction = 'wide',
                   idvar = c('id', 'sex'), # column(s) that determine which rows should be grouped together in the wide data
                   v.names = 'val', # column containing values to widen
                   timevar = 'week', # column from which resulting wide column names are pulled
                   sep = '_') # the `sep` argument allows a user to specify how the contents of `timevar` should be joined with the name of the `v.names` variable to form wide column names
df_wide
```

The **tidyverse** function for taking data from long to wide is `pivot_wider()`. To widen `df_long` using `pivot_longer()`, a user would write:

```{r}
library(tidyverse)
df_wide_PW <- pivot_wider(df_long,
                          id_cols = c('id', 'sex'),
                          values_from = 'val',
                          names_from = 'week',
                          names_prefix = 'week_') # `names_prefix` specifies a string to paste in front of the contents of 'week' in the resulting wide column names
df_wide_PW
```

`pivot_wider()` offers a lot of usability when widening relatively complicated long data structures. For example, say we want to widen both of the long versions of the animals data frame created above.

To widen the version of the long data that has a column for each of the measures (playfulness and excitement):

```{r}
animals_long_1
animals_wide <- pivot_wider(animals_long_1,
                            id_cols = c('animal', 'lives_in_water'),
                            values_from = c('jan', 'feb'),
                            names_from = 'measure',
                            names_sep = '_')
animals_wide
```

To widen the version of the long data that has one column containing all the values of playfulness and excitement together:

```{r}
animals_long_2
animals_wide <- pivot_wider(animals_long_2,
                            id_cols = c('animal', 'lives_in_water'),
                            values_from = 'val',
                            names_from = c('month', 'measure'),
                            names_sep = '_')
animals_wide
```

## Merge/Join

### Left Join

#### Python {-}
#### R {-}

### Right Join

#### Python {-}
#### R {-}

### Inner Join

#### Python {-}
#### R {-}

### Outer Join

#### Python {-}
#### R {-}