# Aggregation and Group Operations

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at manipulating and summarizing data by groups.

## Cross tabulation

Cross tabulation is the process of determining frequencies per group (or values based on frequencies, like proportions), with groups delineated by one or more variables (e.g., nationality and sex).

The Python and R examples of cross tabulation below both make use of the following dataset, `dat`:

```{r, echo = F}
set.seed(100)
dat <- data.frame(nationality = c('Canadian', 'French', 'French', 'Egyptian', 'Canadian'),
                  sex = c('m', 'f', 'f', 'm', 'f'))
```
```{python, echo = F}
dat = r.dat
```
```{r}
dat
```

#### Python {-}

The **pandas** package contains a `crosstab()` function for cross tabulation with two or more variables. The `groupby()`, also in **pandas**, facilitates cross tabulation by one or more variables when used in combination with `count()`.

```{python}
import pandas as pd
pd.crosstab(dat.nationality, dat.sex)
dat.groupby(by = 'nationality').nationality.count()
dat.groupby(by = ['nationality', 'sex']).nationality.count()
# Or: dat.groupby(by = ['nationality', 'sex']).sex.count()
```

#### R {-}

The `table()` function performs cross tabulation in R. A user can enter a single grouping variable or enter multiple grouping variables separated by a comma(s). The `xtabs()` function also computes cross-tabs; a user enters the variables to be used for grouping in formula notation.

```{r}
table(dat$nationality)
table(dat$nationality, dat$sex)
xtabs(formula = ~nationality + sex, data = dat)
```

## Group summaries

Computing statistical summaries per group.

#### Python {-}


#### R {-}

