# Aggregation and Group Operations

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at manipulating and summarizing data by groups.

## Cross tabulation

Cross tabulation is the process of determining frequencies per group (or determining values based on frequencies, like proportions), with groups delineated by one or more variables (e.g., nationality and sex).

The Python and R examples of cross tabulation below both make use of the following dataset, `dat`:

```{r, echo = F}
set.seed(100)
dat <- data.frame(nationality = c('Canadian', 'French', 
                                  'French', 'Egyptian', 
                                  'Canadian'),
                  sex = c('m', 'f', 'f', 'm', 'f'))
```
```{python, echo = F}
dat = r.dat
```
```{r}
dat
```

#### Python {-}

The **pandas** package contains a `crosstab()` function for cross tabulation with two or more variables. Alternatively, the `groupby()` function, also in **pandas**, facilitates cross tabulation by one or more variables when used in combination with `count()`.

```{python}
import pandas as pd
pd.crosstab(dat.nationality, dat.sex)
dat.groupby(by = 'nationality').nationality.count()
dat.groupby(by = ['nationality', 'sex']).nationality.count()
# Or: dat.groupby(by = ['nationality', 'sex']).sex.count()
```

#### R {-}

The `table()` function performs cross tabulation in R. A user can enter a single grouping variable or enter multiple grouping variables separated by a comma(s). The `xtabs()` function also computes cross-tabs; a user enters the variables to be used for grouping in formula notation.

```{r}
table(dat$nationality)
table(dat$nationality, dat$sex)
xtabs(formula = ~nationality + sex, data = dat)
```

## Group summaries

Computing statistical summaries per group.
```{r echo=FALSE}
rm(mtcars)
data("mtcars")
```
```{python echo=FALSE}
mtcars = r.mtcars
```

#### Python {-}

The `groupby()` function from **Pandas** splits up a data set based on one or more grouping variables. Summarizing functions---like `mean()`, `sum()`, and so on---can then be applied to those groups. In the first example below, we use `groupby()` to group rows of the `mtcars` dataset by the number of cylinders each car has; from there, we select just the `mpg` column and call `mean()`, thus producing the average miles per gallon within each cylinder group. In the second example, we again group observations by `cyl`, but instead of then selecting just the `mpg` column, we directly call `mean()`; this gives the mean for each variable in the data set within each cylinder group. Finally, in the third example, we group by two variables---`cyl` and `vs`---and then use the `describe()` function to generate a set of descriptive statistics for `mpg` within each `cylinder`*`vs` group (e.g., mean, SD, minimum, etc.).

```{python}
import pandas as pd

mean_mpg_by_cyl = mtcars.groupby(by = 'cyl')['mpg'].mean()
print(mean_mpg_by_cyl)

means_all_vars = mtcars.groupby(by = 'cyl').mean()
print(means_all_vars)

mpg_by_cyl_vs = mtcars.groupby(by = ['cyl', 'vs'])['mpg'].describe()
print(mpg_by_cyl_vs)
```

#### R {-}

The `aggregate()` function can be used to generate by-group statistical summaries based on one or more grouping variables. Grouping variables can be declared as a list in the function's `by` argument. Alternatively, the grouping variable(s) and the variable to be summarized can be passed to `aggregate()` in formula notation: `var_to_be_aggregated ~ grouping_var_1 + ... + grouping_var_N`. The summarizing function (e.g., `mean()`; `median()`; etc.) is declared in the `FUN` argument.

```{r agg1}
# One grouping variable
# Calculating mean of `mpg` in each `cyl` group
aggregate(x = mtcars$mpg, 
          by = list(cyl = mtcars$cyl), 
          FUN = "mean") 
```

Adding `drop=FALSE` ensures all combinations of levels are returned even if no data exist at that combination. Below the final row is `NA` since there are no 8-cylinder cars with a "straight" engine (vs = 1).

```{r agg2}
# Two or more grouping variables
# Calculating max of `mpg` in each `cyl`*`vs` group
aggregate(x = mtcars$mpg, 
          by = list(cyl = mtcars$cyl, vs = mtcars$vs), 
          FUN = "max", drop = FALSE) 
```

```{r, eval = F}
# Or, specify the variable to summarize and the grouping variables in formula notation
aggregate(mpg ~ cyl + vs, data = mtcars, FUN = max)
```

The **tidyverse** also offers a summarizing function, `summarize()` (or `summarise()`, for the Britons), which is in the **dplyr** package. After grouping a data frame/tibble (with, e.g., **dplyr**'s `group_by()` function), a user passes it to `summarize()`, specifying in the function call how the summary statistic should be calculated.

```{r}
library(dplyr)
mtcars %>% 
  group_by(cyl, vs) %>% 
  summarize(avg_mpg = mean(mpg))
```

`summarize()` makes it easy to specify relatively complicated summary calculations without needing to write an external function.

```{r}
mtcars %>% 
  group_by(cyl, vs) %>% 
  summarize(avg_mpg = mean(mpg),
            complicated_summary_calculation = 
              min(mpg)^0.5 * 
              mean(wt)^0.5 + 
              mean(disp)^(1/mean(hp)))
```

## Centering and Scaling

_Centering_ refers to subtracting a constant, such as the mean, from every one of set of values. This is sometimes performed to aid interpretation of linear model coefficients.

_Scaling_ refers to rescaling a column or vector of values such that their mean is zero and their standard deviation is one. This is sometimes performed to put multiple variables on the same scale and is often recommended for procedures such as principal components analysis (PCA).

#### Python {-}

The `scale()` function from the **preprocessing** module of the **scikit-learn** package provides one-step centering and scaling. To center a variable at zero without scaling it, use `scale()` with `with_mean = True` and `with_std = False` (both are `True` by default).

```{python}
from sklearn import preprocessing

centered_mpg = preprocessing.scale(mtcars.mpg, with_mean = True, with_std = False)
centered_mpg.mean()
```

To scale a variable after centering it (so that its mean is zero and its standard deviation is one), use `scale()` with `with_mean = True` and `with_std = True`.

```{python}
from sklearn import preprocessing

scaled_mpg = preprocessing.scale(mtcars.mpg, with_mean = True, with_std = True)
scaled_mpg.mean()
scaled_mpg.std()
```

#### R {-}

The `scale()` function can both center and scale variables.

To center a variable without scaling it, call `scale()` with the `center` argument set to `TRUE` and the `scale` argument set to `FALSE`. The variable's mean will be subtracted off of each of the variable values. (Note: If desired, the `center` argument can be set to a numeric value instead of `TRUE`/`FALSE`; in that case, each variable value will have the argument value subtracted off of it.)

```{r}
centered_mpg <- scale(mtcars$mpg, center = T, scale = F)
mean(centered_mpg)
```

To scale a variable (while also centering it), call `scale()` with the `center` and `scale` arguments set to `TRUE` (these are the default argument values). The variable's mean will be subtracted off of each of the variable values, and each value will then be divided by the variable's standard deviation. (Note: As with the `center` argument, the `scale` argument can also be set to a numeric value instead of `TRUE`/`FALSE`; in that case, the divisor will be the argument value instead of the standard deviation.)

```{r}
scaled_mpg <- scale(mtcars$mpg, center = T, scale = T)
mean(scaled_mpg)
sd(scaled_mpg)
```