# Aggregation and Group Operations

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at manipulating and summarizing data by groups.

## Cross tabulation

Cross tabulation is the process of determining frequencies per group (or determining values based on frequencies, like proportions), with groups delineated by one or more variables (e.g., nationality and sex).

The Python and R examples of cross tabulation below both make use of the following dataset, `dat`:

```{r, echo = F}
set.seed(100)
dat <- data.frame(nationality = c('Canadian', 'French', 
                                  'French', 'Egyptian', 
                                  'Canadian'),
                  sex = c('m', 'f', 'f', 'm', 'f'))
```

```{python, echo = F}
dat = r.dat
```

```{r}
dat
```

#### Python {-}

The **pandas** package contains a `crosstab()` function for cross tabulation with two or more variables. Alternatively, the `groupby()` function, also in **pandas**, facilitates cross tabulation by one or more variables when used in combination with `count()`.

```{python}
import pandas as pd
pd.crosstab(dat.nationality, dat.sex)
dat.groupby(by = 'nationality').nationality.count()
dat.groupby(by = ['nationality', 'sex']).nationality.count()
# Or: dat.groupby(by = ['nationality', 'sex']).sex.count()
```

#### R {-}

The `table()` function performs cross tabulation in R. A user can enter a single grouping variable or enter multiple grouping variables separated by a comma(s). The `xtabs()` function also computes cross-tabs; a user enters the variables to be used for grouping in formula notation.

```{r}
table(dat$nationality)
table(dat$nationality, dat$sex)
xtabs(formula = ~nationality + sex, data = dat)
```

## Group summaries

Computing statistical summaries per group.

#### Python {-}

#### R {-}

The `aggregate()` function allows a user to easily generate by-group statistical summaries based on one or more grouping variables. Grouping variables can be declared as a list in the function's `by` argument. Alternatively, the grouping variables and the variable to be summarized can be passed to `aggregate()` in formula notation: `var_to_be_aggregated ~ grouping_var_1 + ... + grouping_var_N`. The summarizing function (e.g., `mean()`; `median()`; etc.) is declared in the `FUN` argument.

```{r echo=FALSE}
rm(mtcars)
data("mtcars")
```


```{r agg1}
# One grouping variable
# Calculating mean of `mpg` in each `cyl` group
aggregate(x = mtcars$mpg, 
          by = list(cyl = mtcars$cyl), 
          FUN = "mean") 
```

Adding `drop=FALSE` ensures all combinations of levels are returned even if no data exist at that combination. Below the final row is `NA` since there are no 8 cylinder cars with a "straight" engine (vs = 1).

```{r agg2}
# Two or more grouping variables
# Calculating max of `mpg` in each `cyl`*`vs` group
aggregate(x = mtcars$mpg, 
          by = list(cyl = mtcars$cyl, vs = mtcars$vs), 
          FUN = "max", drop = FALSE) 
```


```{r, eval = F}
# Or, specify the variable to summarize and the grouping variables in formula notation
aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
aggregate(mpg ~ cyl + vs, data = mtcars, FUN = max)
```

The **tidyverse** also offers a summarizing function, `summarize()` (or `summarise()`, for the Britons), which is in the **dplyr** package. After grouping a data frame/tibble (with, e.g., **dplyr**'s `group_by()` function), a user passes it to `summarize()`, specifying in the function call how the summary statistic should be calculated.

```{r}
library(dplyr)
mtcars %>% 
  group_by(cyl, vs) %>% 
  summarize(avg_mpg = mean(mpg))
```

`summarize()` makes it easy to specify relatively complicated summary calculations without needing to write an external function.

```{r}
mtcars %>% 
  group_by(cyl, vs) %>% 
  summarize(avg_mpg = mean(mpg),
            complicated_summary_calculation = 
              min(mpg)^0.5 * 
              mean(wt)^0.5 + 
              mean(disp)^(1/mean(hp)))
```

## Centering and Scaling

_Centering_ refers to subtracting a constant from every value, such as the mean. This is sometimes performed to aid interpretation of linear model coefficients. 

_Scaling_ refers to rescaling a column or vector of values such that their mean is 0 and their standard deviation is 1. This is sometimes performed to put multiple variables on the same scale and is often recommended for procedures such as principal components analysis (PCA).

#### Python {-}

#### R {-}

The `scale()` function can both center and scale variables.

To center a variable without scaling it, call `scale()` with the `center` argument set to `TRUE` and the `scale` argument set to `FALSE`. The variable's mean will be subtracted off of each of the variable values. (Note: If desired, the `center` argument can be set to a numeric value instead of `TRUE`/`FALSE`; in that case, each variable value will have the argument value subtracted off of it.)

```{r}
centered_mpg <- scale(mtcars$mpg, center = T, scale = F)
mean(centered_mpg)
```

To scale a variable (while also centering it), call `scale()` with the `center` and `scale` arguments set to `TRUE` (these are the default argument values). The variable's mean will be subtracted off of each of the variable values, and each value will then be divided by the variable's standard deviation. (Note: As with the `center` argument, the `scale` argument can also be set to a numeric value instead of `TRUE`/`FALSE`; in that case, the divisor will be the argument value instead of the standard deviation.)

```{r}
scaled_mpg <- scale(mtcars$mpg, center = T, scale = T)
mean(scaled_mpg)
sd(scaled_mpg)
```