# Basic Plotting and Visualization

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = '', prompt = FALSE, collapse = FALSE)
```

This chapter discusses methods for creating plots to explore and understand data. Visualization in Python and R is a gigantic and evolving topic; we don't pretend to present a comprehensive comparison. 

The plots below make use of the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data set, which contains various measurements for 344 penguins across three islands in the Antarctic Palmer Archipelago. The data were collected by [Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and colleagues, and they were made available under a [CC0 public domain license](https://creativecommons.org/share-your-work/public-domain/cc0/) by Allison Horst, Alison Hill, and Kristen Gorman.

For the R sections below, we show how to make each plot with base R and with **ggplot2**.

```{r, echo = F}
library(palmerpenguins)
```

```{python, echo = F}
penguins = r.penguins
```

Here's a glimpse at the data set:
```{r}
head(penguins)
```

## Histograms

Visualizing the distribution of numeric data.

#### Python {-}

The Python library **Matplotlib** provides the `hist()` function for plotting a histogram. There are many arguments that can be specified in the `hist()` function for customizing the histogram plot to fit your needs: Some parameters include the number of bins, the upper and lower bounds on each bin, the coloration, and more.

Below, we show how to generate a histogram of the bill lengths from the `penguins` dataset. We specified 30 bins, each of which is light blue with a black outline of `linewidth = 1`. The `hist()` function defaults to showing no bin outline, which can make it difficult to distinguish bins clearly, so we add in the outlines here. 

Note that the bins are left-inclusive and right-exclusive. For example, if a particular bin spans the range of 1 to 3, the bin will include the value 1 but will exclude the value 3 (and will include all values between 1 and 3). In short, bin ranges are [x1,x2), where x1 is the starting point of the bin and x2 is the ending point of the bin. 

Notice the semicolon at the end of the `plt.hist()` function. This suppresses the printing of the array generated to create the histogram. 

```{python}
import matplotlib.pyplot as plt

plt.clf() # clears the current figure to ensure that multiple plots are not overlaid
plt.hist(penguins.bill_length_mm, bins = 30, range = (30, 60),
         color = 'lightblue', edgecolor = 'k', linewidth = 1);
plt.title("Penguin Bill Lengths")
plt.xlabel("Bill Length (mm)")
plt.ylabel("Count")
plt.show()
```

#### R {-}

Base R's `hist()` function generates histograms, and features of the histogram---like the bar color, number of bins/breaks, and so on---can be easily customized with arguments as demonstrated below.

```{r}
hist(penguins$bill_length_mm, breaks = 25, col = 'lightblue', 
     xlim = c(30, 60),
     main = 'Penguin Bill Lengths', 
     xlab = 'Bill Length (mm)', ylab = 'Count')
```

The **ggplot2** method for generating histograms follows the standard **ggplot2** syntax: Initialize a plot with `ggplot()`, then add layers thereto. Here, the layer to add is `geom_histogram()`.

```{r, warning = F}
library(ggplot2)

ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(fill = 'lightblue', color = 'black', bins = 25) +
  xlim(30, 60) + 
  labs(title = 'Penguin Bill Lengths',
       x = 'Bill Length (mm)', y = 'Count')
```

## Barplots

Visualizing the distribution of categorical data.

#### Python {-}

For this example, we generate a barplot showing how many penguins of each species---Adelie, Chinstrap, Gentoo---are represented in the data. We show two ways of doing so here. 

First, we can use the function `bar()` from the **Matplotlib** plotting library. We determine the per-species penguin counts and then use that data to create the bar plot. 

```{python}
import matplotlib.pyplot as plt 

# Determine the number of each species
adelie_counts = len(penguins.loc[penguins["species"] == "Adelie"])
chinstrap_counts = len(penguins.loc[penguins["species"] == "Chinstrap"])
gentoo_counts = len(penguins.loc[penguins["species"] == "Gentoo"])

# Save the counts information as an arrays for input into the bar() function
spec = ["Adelie", "Chinstrap", "Gentoo"]
counts = [adelie_counts, chinstrap_counts, gentoo_counts]

plt.clf()
plt.bar(spec, counts)
plt.show()
```

The data are stored in a **pandas** Dataframe, which has its own built-in plotting module, `plot`. As an alternative approach to the one above, we can use the **pandas** `bar()` function to recreate the barplot.

```{python}
plt.clf()
penguins["species"].value_counts().plot.bar()
plt.show()
```

Using this second approach, we generated the same barplot with far less effort. Using the built-in **pandas** plotting routine proved to be the more efficient method here. 

#### R {-}

To form barplots, we'll first create a summary data frame containing the statistics we're looking to plot. Here, that's simply the sample size of each species in the data set.

```{r}
species_counts <- as.data.frame(xtabs(~ species, data = penguins))
species_counts
```

We can plot those values using the `barplot()` function in base R, specifying arguments along the way to customize the title/axis labeling, bar colors, and range of the y axis. To add values above the bars, we can follow `barplot()` with a `text()` call as below.

```{r}
penguin_plot <- barplot(Freq ~ species, data = species_counts, 
                        col = c('lightblue', 'cornflowerblue', 'darkslateblue'),
                        main = 'Species Sample Size', 
                        xlab = 'Species', ylab = 'Count', ylim = c(0, 200))
text(x = penguin_plot, y = species_counts$Freq + 10, 
     labels = species_counts$Freq)
```

To recreate the bar plot above with **ggplot2**, one can add a `geom_bar()` layer to a plot initialized with `ggplot()`.

```{r, warning = F}
ggplot(species_counts, aes(x = species, y = Freq)) +
  geom_bar(aes(fill = species), stat = 'identity') +
  
  # scale_fill_manual() is used here for bar-color customization
  scale_fill_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) + 
  labs(title = 'Species Sample Size', x = 'Species', y = 'Count') +
  
  # For simplicity, we omit the legend 
  theme(legend.position = 'none') + 
  ylim(0, 200) + 
  
  # geom_text() is used here to add counts above the bars
  geom_text(aes(label = Freq, vjust = -0.5)) 
```

## Scatterplot

Visualizing the relationship between two numeric variables.

#### Python {-}

The `scatter()` function, part of **Matplotlib**, can produce scatterplots in Python. The `x` and `y` arguments specify the points to plot. In the example below, we also use the `c` and `marker` arguments to customize the point color and point shape, respectively. The `xlabel()`, `ylabel()`, and `title()` functions customize plot labels.

```{python}
import matplotlib.pyplot as plt

# Remove NA rows, as we only want to plot present data
penguins_no_na = penguins.dropna() 

plt.clf()

# 'd' generates diamond markers (learn more about available marker shapes here: https://matplotlib.org/stable/api/markers_api.html#module-matplotlib.markers)
plt.scatter(x = penguins_no_na['body_mass_g'], 
            y = penguins_no_na['flipper_length_mm'], 
            c = 'lightblue', marker = 'd') 
plt.xlabel('Body Mass (g)')
plt.ylabel('Flipper Length (mm)')
plt.title('Scatterplot of Body Mass and Flipper Length')

plt.show()
```

#### R {-}

Scatterplots can be generated in base R with the `plot()` function. The `pch` argument below modifies the point shape (e.g., 20 = solid circle; 24 = unfilled triangle; etc.)

```{r}
plot(x = penguins$body_mass_g, y = penguins$flipper_length_mm, 
     col = 'navy', pch = 20,
     main = 'Scatterplot of Body Mass and Flipper Length', 
     xlab = 'Body Mass (g)', ylab = 'Flipper Length (mm)')
```

To generate a scatterplot with **ggplot2**, initialize a plot with `ggplot()`, then add a layer of points with `geom_point()`.

```{r, warning = F}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(color = 'navy') +
  labs(title = 'Scatterplot of Body Mass and Flipper Length', 
       x = 'Body Mass (g)', y = 'Flipper Length (mm)')
```

## Stripcharts

Stripcharts, or strip plots, are one-dimensional scatterplots. Like boxplots, they reveal the distribution of a numeric variable within levels of a categorical variable.

#### Python {-}

The **seaborn** package provides the `stripplot()` function. The user simply specifies which variables they want on the x and y axes. Below, we specify `island` on the y axis to see the distribution of `bill_depth_mm` horizontally. Specify the **pandas** DataFrame using the `data` argument. Finally, create the plot using `plt.show()`.

```{python}
import seaborn as sns
import matplotlib.pyplot as plt
sns.stripplot(x = "bill_depth_mm", y = "island", data = penguins)
plt.show()
```

The [stripplot](https://seaborn.pydata.org/generated/seaborn.stripplot.html) help page provides more examples. 

#### R {-}

Base R offers the `stripchart()` function. To indicate the numeric variable and the grouping variable, you can use formula notation: `numeric_var ~ grouping_var`. Adding `method = 'jitter'` to the arguments spreads the points out slightly within each level of the grouping variable, making it easier to see points that might otherwise be obscured by overlap.

```{r}
stripchart(bill_depth_mm ~ island, data = penguins, 
           method = 'jitter',
           ylab = 'Island', xlab = 'Bill Depth (mm)', 
           main = 'Stripchart of Bill Depth by Island')
```

Stripcharts can also be made with **ggplot2**'s  `geom_jitter()` function, as shown below. You can control the amount of jitter with a `position` argument in `geom_jitter()`.

```{r}
ggplot(penguins, aes(x = island, y = bill_depth_mm)) +
  geom_jitter(aes(color = island), position = position_jitter(0.1)) +
  
  # scale_fill_manual() is used to manually specify group colors 
  # once aes(color = island) is specified in `geom_jitter()`
  scale_color_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) +
  
  labs(title = 'Stripchart of Bill Depth by Island', 
       x = 'Island', y = 'Bill Depth (mm)') +
  theme(legend.position = 'none')
```

## Boxplots

Visualizing the relationship between a numeric variable and a categorical variable via five-number summaries.

#### Python {-}

The `boxplot()` function in **seaborn** generates boxplots, and functions from **Matplotlib**, like `xlabel()` and `ylabel()`, can be used to control the aesthetics of the plot.

```{python}
import seaborn as sns
import matplotlib.pyplot as plt
plt.figure()
sns.boxplot(x = "island", y = "bill_depth_mm", data = penguins)
plt.xlabel("Island")
plt.ylabel("Bill Depth (mm)")
plt.title("Bill Depth by Island")
plt.show()
```

#### R {-}

The `boxplot()` function in base R generates boxplots. A user specifies the grouping variable and the numeric variable to be plotted in formula notation: `y ~ grouping_var`.

```{r}
boxplot(bill_depth_mm ~ island, data = penguins, 
        col = c('lightblue', 'cornflowerblue', 'darkslateblue'),
        main = 'Boxplot of Bill Depth by Island', 
        xlab = 'Island', ylab = 'Bill Depth (mm)')
```

To generate a boxplot with **ggplot2**, add a `geom_boxplot()` layer to a plot initialized with `ggplot()`.

```{r, warning = F}
ggplot(penguins, aes(x = island, y = bill_depth_mm)) +
  geom_boxplot(aes(fill = island)) +
  scale_fill_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) +
  labs(title = 'Boxplot of Bill Depth by Island', 
       x = 'Island', y = 'Bill Depth (mm)') +
  theme(legend.position = 'none')
```

## Facet plots

Facet plots (also called trellis plots, lattice plots, and conditional plots) are comprised of multiple smaller plots, where each subplot contains a subset of the overall data, with subsets defined by one or more faceting variables.

#### Python {-}

The **seaborn** package provides several functions for creating facet plots, including `relplot()`, `displot()`, `catplot()`, and `lmplot()`. Below we demonstrate the `lmplot()` function, which allows a user to create scatterplots at certain levels of categorical variable. A user specifies the x and y variables as character strings using the `x` and `y` arguments, respectively. The grouping variable is identified with either the `col` or `row` arguments. By default, a least-squares lines is added to each plot. Setting `ci = None` suppresses the confidence interval ribbon.

```{python}
import seaborn as sns
plt.clf()
sns.lmplot(x = "bill_length_mm", y = "bill_depth_mm", 
           col = "species",
           data = penguins, ci = None)
```

To facet by two variables, provide variables to both the `col` and `row` arguments.

```{python}
plt.clf()
sns.lmplot(x = "bill_length_mm", y = "bill_depth_mm", 
           col = "species", row = "sex",
           data = penguins, ci = None)
```

Set `lowess = True` for smooth trend lines. Notice also that color can be mapped to the same variable used for faceting.

```{python}
plt.clf()
sns.lmplot(x = "bill_length_mm", y = "bill_depth_mm", 
           col = "species", hue = "species",
           data = penguins, ci = None, lowess = True)
```


The [1lmplot()1](https://seaborn.pydata.org/generated/seaborn.lmplot.html#seaborn.lmplot) help page showcases other examples.

#### R {-}

The `coplot()` function in base R produces conditioning plots using formula notation: `y ~ x | grouping_var`. The `rows` and `columns` arguments control layout. Below we specify one row of plots. The `panel` argument controls what action is carried out in each plot. The default is a scatterplot. Below we use the base R `panel.smooth` function to create scatter plots with a smooth trend line.

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(bill_depth_mm ~ bill_length_mm | species, 
       data = penguins,
       panel = panel.smooth,
       rows = 1, col = "navy")
```

To condition on two variables, use formula notation with syntax: `y ~ x | grp_var1 * grp_var2`.

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(bill_depth_mm ~ bill_length_mm | species * sex, 
       data = penguins,
       panel = panel.smooth,
       col = "navy")
```

The labels of the conditioning variables unfortunately use a lot of real estate in the margins, and there is no easy way to modify that. However, this design works quite well when we condition on a _numeric variable_. The `coplot()` function automatically creates overlapping group intervals to condition on, and the stacked layout of the labels helps us visualize how the relationship between x and y changes between the groups. To manually set the number of groups, use the `number` argument. Below, we specify four groups to be generated for `bill_length_mm`.

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(flipper_length_mm ~ body_mass_g | bill_length_mm, 
       data = penguins,
       panel = panel.smooth,
       number = 4, 
       rows = 1, col = "navy")
```

Alternatively, **ggplot2** provides a intuitive and easy-to-use method for generating facet plots: A user specifies the aesthetics of the plot using standard **ggplot2** syntax (i.e., as a series of added layers) and then adds an additional call, `facet_wrap()` or `facet_grid()` (the differences are discussed below), specifying the faceting variable(s) to split up the plots by.

```{r, warning = F}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  
  # Add least-squares line
  geom_smooth(method = 'lm', se = FALSE, color = 'salmon') + 
  
  # Use formula notation, a character vector, or vars() to 
  # specify faceting variables; e.g., ~species, c('species'), 
  # or vars(species)
  facet_wrap(~ species)
```

The number of rows and columns can be manually specified with `nrow` and `ncol` arguments in `facet_wrap()`. By default, the x and y axes of all facet plots will be on the same scale. The axis ranges can be set to vary freely by adding `scales = 'free'` as an argument (or, alternatively, `scales = 'free_x'` or `scales = 'free_y'` to free just the x or y axis).

Both `facet_wrap()` and `facet_grid()` can be used to make facet plots. When faceting based on multiple variables (e.g., species and sex), `facet_wrap()` will drop group combinations for which there are no data points, whereas `facet_grid()` will generate a plot for all possible group combinations:

```{r, warning = F}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  geom_smooth(method = 'lm', se = F, color = 'salmon') +
  facet_wrap(vars(species, sex))

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  geom_smooth(method = 'lm', se = F, color = 'salmon') +
  
  # Note that facet_grid() has separate `rows` and `cols` arguments 
  # for specifying faceting variables
  facet_grid(rows = vars(sex), cols = vars(species)) 
```
