# Basic Plotting and Visualization

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

This chapter looks at creating basic plots to explore and understand data. Visualization in Python and R is a gigantic and evolving topic. We don't pretend to present a comprehensive comparison. 

The plots below make use of the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data set, which contains various measurements for 344 penguins across three islands in the Antarctic Palmer Archipelago. The data were collected by [Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and colleagues, and they were made available under a [CC0 public domain license](https://creativecommons.org/share-your-work/public-domain/cc0/) by Allison Horst, Alison Hill, and Kristen Gorman.

For the R sections below, we discuss how to generate plots using base R and using **ggplot2**.

```{r, echo = F}
library(palmerpenguins)
```
```{python, echo = F}
penguins = r.penguins
```

Here's a glimpse at the data set:
```{r}
head(penguins)
```

## Histograms

Visualizing the distribution of numeric data.

#### Python {-}

The Python plotting library Matplotlib's `hist()` function computes and plots a histogram. There are many parameters that can be specified within the `hist()` function so that you can customize the output histogram plot to best fit your needs. Some parameters include the number of bins, the upper and lower bounds on each bin, weights, colors, and more. 

Below we show a histogram of the bill length from the dataset. We specified 30 bins each of which is light blue with a black outline of linewidth 1. The `hist()` defaults to no outline which can make it difficult to distinguish bins clearly, so we add in the bin outlines here. 

One thing to note is that the bins are left inclusive and right exclusive. For example, if a particular bin spans the range of 1 to 3, the bin will include the value 1 but will exclude the value 2 (and will include all values between 1 and 3). In short, bin ranges are as follows [x1,x2) where x1 is the starting point of the bin and x2 is the ending point of the bin. 

Notice the semicolon at the end of the `plt.hist()` function. This suppresses the printing of the array generated to create the histogram. 

```{python}
import matplotlib.pyplot as plt

plt.clf()
plt.hist(penguins.bill_length_mm, bins=30, range=(30,60),
         color='lightblue', edgecolor='k', linewidth=1);
plt.title("Penguin Bill Lengths")
plt.xlabel("Bill Length (mm)")
plt.ylabel("Count")
plt.show()
```

#### R {-}
Base R's `hist()` function generates histograms, and features of the histogram---like the bar color, number of bins/breaks, and so on---can be easily customized as below.
```{r}
hist(penguins$bill_length_mm, breaks = 25, col = 'lightblue', xlim = c(30, 60),
     main = 'Penguin Bill Lengths', xlab = 'Bill Length (mm)', ylab = 'Count')
```

The **ggplot2** method for generating histograms follows the standard **ggplot2** syntax: Initialize a plot with `ggplot()`, and then add layers thereto, specifying aesthetic properties along the way. Here, the layer to add is `geom_histogram()`.
```{r, warning = F}
ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(fill = 'lightblue', color = 'black', bins = 25) +
  xlim(30, 60) + labs(title = 'Penguin Bill Lengths', x = 'Bill Length (mm)', y = 'Count')
```

## Barplots

Visualizing the distribution of categorical data.

#### Python {-}
For this example, we will generate a bar plot showing how many of each species - Adelie, Chinstrap, Gentoo - we have in our dataset. We go through two ways of doing this here. 

First, we use the Matplotlib plotting library to create the bar plot using the function `bar()`. To start, we determine the number of each species, then use that data to create the bar plot. 

```{python}
import matplotlib.pyplot as plt 

# Determine the number of each species
adelie_counts = len(penguins.loc[penguins["species"]=="Adelie"])
chinstrap_counts = len(penguins.loc[penguins["species"]=="Chinstrap"])
gentoo_counts = len(penguins.loc[penguins["species"]=="Gentoo"])

# Save the counts information into arrays to be inputted into the bar() function
spec = ["Adelie","Chinstrap","Gentoo"]
counts = [adelie_counts,chinstrap_counts,gentoo_counts]

plt.clf() # clears the figure to ensure that multiple plots are not overlaid
plt.bar(spec,counts)
plt.show()
```

Our data is stored in a `pandas Dataframe`, which has it's own built-in plotting module, `plot`. Here we create the same bar plot by using the pandas `bar()` function. 

```{python}
plt.clf()
penguins["species"].value_counts().plot.bar()
plt.show()
```
One thing to note here is that we generated the same bar plot containing the same information with way less effort. Using the built-in pandas plotting routine proved to be the more efficient method here. 

#### R {-}
To form barplots, we'll first take the **penguins** data set and create a summary data frame containing the statistics we're looking to plot. Here, that's simply the sample size of each species in the data set.
```{r}
species_counts <- as.data.frame(xtabs(~ species, data = penguins))
species_counts
```

We can plot those values using the `barplot()` function in base R, specifying arguments along the way to customize the title/axis labeling, bar colors, and range of the y axis. To add values above the bars, we can follow `barplot()` with a `text()` call as below.
```{r}
penguin_plot <- barplot(Freq ~ species, data = species_counts, col = c('lightblue', 'cornflowerblue', 'darkslateblue'),
                        main = 'Species Sample Size', 
                        xlab = 'Species', ylab = 'Count', ylim = c(0, 200))
text(x = penguin_plot, y = species_counts$Freq + 10, 
     labels = species_counts$Freq)
```

To recreate the barplot above with **ggplot2**, one can add a `geom_bar()` layer to a plot initialized with `ggplot()`.
```{r, warning = F}
ggplot(species_counts, aes(x = species, y = Freq)) +
  geom_bar(aes(fill = species), stat = 'identity') +
  scale_fill_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) + # scale_fill_manual() is used here for bar-color customization
  labs(title = 'Species Sample Size', x = 'Species', y = 'Count') +
  theme(legend.position = 'none') + ylim(0, 200) + # For simplicity, we omit the legend here using the code at left
  geom_text(aes(label = Freq, vjust = -0.5)) # geom_text() is used here to add counts above the bars
```

## Scatterplot

Visualizing the relationship between two numeric variables.

#### Python {-}

#### R {-}
Scatterplots can be generated in base R with the `plot()` function. The `pch` argument below modifies the point shape (e.g., 20 = solid circle; 24 = unfilled triangle; etc.)
```{r}
plot(penguins$body_mass_g, penguins$flipper_length_mm, 
     col = 'navy', pch = 20,
     main = 'Scatterplot of Body Mass and Flipper Length', 
     xlab = 'Body Mass (g)', ylab = 'Flipper Length (mm)')
```

To generate a scatterplot with **ggplot2**, initialize a plot with `ggplot()`, then add a layer of points with `geom_point()`.
```{r, warning = F}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(color = 'navy') +
  labs(title = 'Scatterplot of Body Mass and Flipper Length', 
       x = 'Body Mass (g)', y = 'Flipper Length (mm)')
```

## Stripcharts

Stripcharts are one-dimensional scatterplots. Like boxplots, they reveal the distribution of a numeric variable "within" levels of a categorical variable.

#### Python {-}

#### R {-}

Base R offers the `stripchart()` function. To indicate the numeric variable and the grouping variable, you can use formula notation: `numeric_var ~ grouping_var`. Adding `methhod = 'jitter'` to the set of arguments spreads the points out slightly within each level of the grouping variable, making it easier to see points that might otherwise be obscured by overlap.
```{r}
stripchart(bill_depth_mm ~ island, data = penguins, method = 'jitter',
           ylab = 'Island', xlab = 'Bill Depth (mm)', main = 'Stripchart of Bill Depth by Island')
```

Stripcharts can also be made with **ggplots2**'s  `geom_jitter()` function, as shown below. You can control the amount of jitter with a `position` argument in `geom_jitter()`.
```{r}
ggplot(penguins, aes(x = island, y = bill_depth_mm)) +
  geom_jitter(aes(color = island), position = position_jitter(0.1)) +
  scale_color_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) + # scale_fill_manual() is used to manually specify group colors once aes(color = island) is specified in `geom_jitter()`
  labs(title = 'Stripchart of Bill Depth by Island', 
       x = 'Island', y = 'Bill Depth (mm)') +
  theme(legend.position = 'none')
```

## Boxplots

Visualizing the relationship between a numeric variable and a categorical variable via five-number summaries.

#### Python {-}

#### R {-}
The `boxplot()` function in base R generates boxplots, and a user specifies the grouping variable and the numeric variable to be plotted in formula notation: `y ~ grouping_var`.
```{r}
boxplot(bill_depth_mm ~ island, data = penguins, col = c('lightblue', 'cornflowerblue', 'darkslateblue'),
        main = 'Boxplot of Bill Depth by Island', xlab = 'Island', ylab = 'Bill Depth (mm)')
```

To generate a boxplot with **ggplot2**, add a `geom_boxplot()` layer to a plot initialized with `ggplot()`.
```{r, warning = F}
ggplot(penguins, aes(x = island, y = bill_depth_mm)) +
  geom_boxplot(aes(fill = island)) +
  scale_fill_manual(values = c('lightblue', 'cornflowerblue', 'darkslateblue')) +
  labs(title = 'Boxplot of Bill Depth by Island', 
       x = 'Island', y = 'Bill Depth (mm)') +
  theme(legend.position = 'none')
```

## Facet plots

Facet plots (also called trellis plots, lattice plots, and conditional plots) are comprised of multiple smaller plots, where each subplot contains a subset of the overall data, with subsets defined by one or more faceting variable.

#### Python {-}

#### R {-}

The `coplot()` function in base R produces conditioning plots using the formula notation: `y ~ x | grouping_var`. The `rows` and `columns` arguments control layout. Below we specify one row of plots. The `panel` argument controls what action is carried out in each plot. The default is a scatterplot. Below we use the base R `smooth.plot` function to create scatter plots with a smooth trend line.

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(bill_depth_mm ~ bill_length_mm | species, 
       data = penguins,
       panel = panel.smooth,
       rows = 1, col = "navy")
```

To condition on two variables, use formula with syntax: `y ~ x | grp_var1 * grp_var2`.

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(bill_depth_mm ~ bill_length_mm | species * sex, 
       data = penguins,
       panel = panel.smooth,
       col = "navy")
```

The labels of the conditioning variables unfortunately use a lot of real estate in the margins and there is no easy way to modify that. However this design works quite well when we condition on a _numeric variable_. The `coplot` function automatically creates overlapping group intervals to condition on, and the stacked layout of the labels helps us visualize how the relationship between y and x changes between the groups. To manually set the number of groups, use the `number` argument. Below we specify 4 groups to be generated for "bill_length_mm".

```{r warning=FALSE, message=FALSE, results='hide'}
coplot(flipper_length_mm ~ body_mass_g | bill_length_mm, 
       data = penguins,
       panel = panel.smooth,
       number = 4, 
       rows = 1, col = "navy")
```


**ggplot2** provides a intuitive and easy-to-use method for generating facet plots: A user specifies the aesthetics of the plot using standard **ggplot2** syntax (i.e., as a series of added layers) and then adds an additional call, `facet_wrap()` (or `facet_grid()`; differences are discussed below), specifying the faceting variable(s) to split up the plots by.
```{r, warning = F}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  geom_smooth(method = 'lm', se = F, color = 'salmon') + # Add basic least-squares line
  facet_wrap(~species) # Use formula notation, a character vector, or vars() to specify faceting variables; e.g., ~species, c('species'), or vars(species)
```

The number of rows and columns can be manually specified with `nrow` and `ncol` arguments in `facet_wrap()`. By default, the x and y axes of all facet plots will be on the same scale. The axis ranges can be set to vary freely by adding `scales = 'free'` as an argument (or, alternatively, `'free_x'` or `'free_y'` to free just the x or y axes.)

Both `facet_wrap()` and `facet_grid()` can be used for faceting. When faceting based on multiple variables (e.g., species and sex), `facet_wrap()` will drop group combinations for which there are no data points, whereas `facet_grid()` will generate a plot for all possible group combinations:
```{r, warning = F}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  geom_smooth(method = 'lm', se = F, color = 'salmon') +
  facet_wrap(vars(species, sex))
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = 'navy') +
  geom_smooth(method = 'lm', se = F, color = 'salmon') +
  facet_grid(rows = vars(sex), cols = vars(species)) # Note that facet_grid() has separate `rows` and `cols` arguments for specifying faceting variables
```